---
title: "Notes on the MSTmap algorithm"
description: >
  How genetic distances are calculated, and what the mvest.bc and detectBadData arguments actually do.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```

> These notes refer to objects built in [Worked example I](worked-example-construction.html). The code below rebuilds them.

```{r preamble, include = FALSE}
sg <- statGen(mapBCu, bychr = FALSE, stat.type = "miss")
mapBC1 <- subset(mapBCu, ind = sg$miss < 1600)
gc <- genClones(mapBC1, tol = 0.95)
cgd <- gc$cgd[-c(1, 4, 5), ]
mapBC2 <- fixClones(mapBC1, cgd, consensus = TRUE)
mm <- statMark(mapBC2, stat.type = "marker")$marker$AB
mapBC3 <- drop.markers(mapBC2, c(markernames(mapBC2)[mm > 0.98],
                                 markernames(mapBC2)[mm < 0.2]))
mapBC3 <- pullCross(mapBC3, type = "missing", pars = list(miss.thresh = 0.1))
mapBC3 <- pullCross(mapBC3, type = "seg.distortion", pars = list(seg.thresh = "bonf"))
mapBC3 <- pullCross(mapBC3, type = "co.located")
mapBC4 <- mstmap(mapBC3, bychr = FALSE, trace = TRUE, dist.fun = "kosambi",
                 p.value = 1e-12)
pg <- profileGen(mapBC4, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
                 id = "Genotype", xo.lambda = 14, layout = c(1, 3), lty = 2,
                 cex = 0.7)
```


This chapter presents miscellaneous additional information associated within the MSTmap algorithm and the arguments supplied to the R/ASMap functions `mstmap.data.frame()` and `mstmap.cross()`.

## MSTmap and distance calculations

After close scrutiny and experience with the MSTmap algorithm it appears there may be circumstances where the algorithm produces inflated genetic distances. This is especially prevalent when the linkage map being constructed contains genotypes with many missing values. In these cases the constructed linkage map is likely to contain runs of missing values for these genotypes. For example, this phenomenon can be seen in the figure below (the dark lines) for the initial constructed linkage map `mapBC4` of the previous chapter. Seven of these genotypes were eventually removed for having excess recombinations across the genome but their missing value structure has also contributed substantially to the inflated distances of the linkage groups.

```{r dist1, fig.cap = "Plot of the missing allele scores for the constructed map mapBC4", fig.alt = "Plot of the missing allele scores for the constructed map mapBC4"}
plotMissing(mapBC4)
```

This problem is easier to explain by presenting various outputs of the linkage map `mapBC4`. Firstly, the map is reconstructed but with `return.imputed = TRUE` added to the call to ensure the imputed marker probability matrix for the linkage groups is added to the returned cross object.

```{r dist3}
mapBC4i <- mstmap(mapBC3, bychr = FALSE, trace = TRUE, dist.fun = "kosambi", p.value = 1e-12, return.imputed = TRUE)
mapBC4i$geno[[1]]$map[1:14]
mapBC4i$imputed.geno[[1]]$map[1:5]
```

To exemplify the problem the markers on L.1 are used. The first 14 markers of L.1 contain the first two co-locating sets of markers and have been calculated to be 0.76 cM apart. The imputed map for L.1 is a reduced version of the linkage map from L.1 that contains all unique markers and one member from each co-locating set. Markers mark246 and mark391 have been chosen as the members for the first two co-locating sets. It can be quickly calculated how many pairs of alleles are observed between the two markers and how many observed recombinations there were across these pairs.

```{r dist4}
len <- apply(mapBC4$geno[[1]]$data[,c(1,5)], 1, function(el)
             length(el[!is.na(el)]))
length(len[len > 1])
bca <- apply(mapBC4i$geno[[1]]$data[,c(1,5)], 1, function(el){
    el <- el[!is.na(el)]
    sum(abs(diff(el)))})
bca[bca > 0]
```

One genuine recombination from 281 observed pairs creates a 0.35 cM distance between the markers, falling short of 0.76 cM. The reason behind this shortfall becomes clear when the imputed marker probability data is printed for the first five markers of the seven genotypes with excessive missing values

```{r dist5}
mapBC4i$imputed.geno[[1]]$data[pg$xo.lambda,1:5]
```

Genotype BC061 has an extended set of missing values across the first four co-located sets of markers. Using the distance formula in the M-step of the EM algorithm given in (5), the estimates of probability for the two missing allele scores in the first two markers for BC061 adds another 1/2 a recombination or $~$`<!-- -->`{=html}0.15 cM distance between the markers. Genotype BC061 will add a similar distance between the second and third marker set and the third and fourth marker sets due the imputed probabilities being close to 0.5 for the run of missing values across the four co-located sets. If there are multiple runs of missing values for genotypes, these small distances accumulate quickly and linkage group lengths appear inflated.

There are two solutions to this problem. The obvious first solution is to remove the genotypes regardless off their usefulness in the genetic map. The second less obvious solution is to use the R/ASMap function `quickEst()` to re-estimate the genetic distances (see section Rapid genetic distance estimation). As the function imputes the missing values in the linkage map with actual allele calls it circumvents the uncertain probabilities that MSTmap uses to form its genetic distances.

```{r dist6}
mapBC4e <- quickEst(mapBC4)
chrlen(mapBC4)
chrlen(mapBC4e)
```

The difference in the lengths of the linkage groups for the two linkage maps is dramatic with an 80+ cM reduction in L.1 and 60+ cM reductions in L.2 and L.4.

In summary, if a linkage map is constructed with genotypes containing many missing values then the missing value structure of the constructed object should be checked. If runs of missing values are detected then the lengths of linkage groups are most likely inflated and estimation of genetic distances should be checked using `quickEst()`.

## Use of the argument `mvest.bc`

The inflated distances caused by excessive missing values across the marker set for a subset of genotypes can be further exacerbated by an injudicious use of the argument `mvest.bc` in the R/ASMap construction functions `mstmap.data.frame()` and `mstmap.cross()`.

To understand how this is possible, there is a requirement to understand how the MSTmap algorithm determines co-segregating or co-locating markers. In the initial stages of the algorithm two markers are deemed co-located if the pairwise distance is zero with this calculation occurring only from pairs of alleles that are observed in both markers. For each co-locating set a marker is chosen as the representative marker and the remaining markers in the set are placed aside with a knowledge of their direct link to the representative marker. The linkage map is then constructed with only the representative markers. This scenario is equivalent to setting `mvest.bc = FALSE`.

If `mvest.bc = TRUE` then missing allele calls contained in each marker are estimated to have probability 0.5 of being an A allele *before* clustering of the markers has occurred. This has an obvious advantage of involving the complete set of genotypes when calculating pairwise information between markers. However, for any two markers where there is some proportion of unobserved allele pairs, the pairwise distance between the markers becomes non-zero and they are deemed to be not co-located. This results in a larger set of markers being involved in the linkage map construction with the original missing allele calls being continually imputed as part of the marker ordering algorithm (see section Marker Ordering). For situations where there is an increased number of missing values across a set of markers for a subset of genotypes, the cumulative distances between adjacent markers quickly increases.

To illustrate this problem, the barley backcross marker set `mapBC3` is constructed using the function `mstmap.cross()` with the argument `mvest.bc = TRUE`.

```{r mvest1}
mapBC4a <- mstmap(mapBC3, bychr = FALSE, trace = TRUE, dist.fun = "kosambi", p.value = 1e-12, mvest.bc = TRUE)
nmar(mapBC4)
```

The clustering algorithm produces an identical number of linkage groups with the same markers within each linkage group. However, the number of uniquely located markers within each linkage group increases dramatically between `mapBC4` and `mapBC4a`. This increase in the number of uniquely located markers also increases the length of each linkage group.

```{r mvest2}
sapply(mapBC4a$geno, function(el) length(unique(round(el$map, 4)))) - sapply(mapBC4$geno, function(el) length(unique(round(el$map, 4))))
chrlen(mapBC4a)
```

Again, the solution to this problem is to use the efficient distance estimation function `quickEst()` to provide an estimate of the genetic distances of the markers in each of the linkage groups.

```{r mvest3}
mapBC4b <- quickEst(mapBC4a)
chrlen(mapBC4b)
```

## Use of the argument `detectBadData`

The last two sections discussed constructed linkage map scenarios where the genetic distance calculations became inflated due to an excessive number of missing values being present for a subset of genotypes in the marker set. If this situation is not present and the linkage group distances are inflated then it is most likely caused by an increased recombination rate in some or all of the genotypes. This is easily checked using the appropriate call to `statGen()` or `profileGen()` after an initial linkage map is constructed. If there is an increased recombination rate in a small subset of genotypes they can be removed before further linkage map construction. If there appears to be a general inflation in the recombination rate across a large set or all of the genotypes then there may be genotyping errors that have occurred in the physical process used to obtain the allele calls for the markers. Consequently, reduction of the linkage group lengths becomes more problematic.

To circumvent this issue, the construction functions `mstmap.data.frame()` and `mstmap.cross()` in the R/ASMap package contain an argument called `detectBadData` that, if set to `TRUE`, instructs MSTmap to detect suspiciously called alleles (see equation 6 of section Marker Ordering for more details). If suspicious allele calls are found they are set to missing and imputed using the EM algorithm detailed in section Marker Ordering. This creates a reduction in the number of recombinations between adjacent markers. The consequence of this is a possibly dramatic reduction in the distances between adjacent markers and overall linkage group lengths.

To exemplify the use of `detectBadData`, simulated genotyping errors are added to the final barley backcross linkage map `mapBC` by randomly switching singular allele calls in the data for each linkage group.

```{r dbd1}
mapBCd <- mapBC
mapBCd$geno <- lapply(mapBCd$geno, function(el){
    ns <- sample(1:ncol(el$data), ncol(el$data)/2, replace = TRUE)
    ns <- cbind(sample(1:nrow(el$data), ncol(el$data)/2, replace = TRUE), ns)
    el$data[ns] <- abs(1 - el$data[ns])
    el$data[el$data == 0] <- 2
    el})
mapBCd <- quickEst(mapBCd)
chrlen(mapBCd)
```

The function `quickEst()` will not ignore the genotyping errors that have been added to the data and linkage groups will appear inflated.

```{r dbd2}
mapBCda <- mstmap(mapBCd, bychr = TRUE, trace = TRUE, dist.fun = "kosambi", p.value = 1e-12, detectBadData = TRUE)
chrlen(mapBCda)
```

A judicious use of the `detectBadData = TRUE` in the `mstmap.cross()` call instructs the algorithm to detect and ignore these errors during the marker ordering algorithm of MSTmap. This deflates the linkage group distances to similar lengths as the linkage groups lengths of the final linkage map `mapBC`.
