---
title: "Notes on the MSTmap algorithm"
description: >
  How genetic distances may become inflated, and what the mvest.bc and
  detectBadData arguments do in practice.
bibliography: ../ref.bib
link-citations: true
nocite: |
  @tb17
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapBC); data(mapBCu)
```

> These notes refer to objects built in
> [Worked example I](worked-example-construction.html). The code below rebuilds
> them.

```{r preamble, include = FALSE}
sg <- statGen(mapBCu, bychr = FALSE, stat.type = "miss")
mapBC1 <- subset(mapBCu, ind = sg$miss < 1600)
gc <- genClones(mapBC1, tol = 0.95)
cgd <- gc$cgd[-c(1, 4, 5), ]
mapBC2 <- fixClones(mapBC1, cgd, consensus = TRUE)
mm <- statMark(mapBC2, stat.type = "marker")$marker$AB
mapBC3 <- drop.markers(mapBC2, c(markernames(mapBC2)[mm > 0.98],
                                 markernames(mapBC2)[mm < 0.2]))
mapBC3 <- pullCross(mapBC3, type = "missing", pars = list(miss.thresh = 0.1))
mapBC3 <- pullCross(mapBC3, type = "seg.distortion", pars = list(seg.thresh = "bonf"))
mapBC3 <- pullCross(mapBC3, type = "co.located")
mapBC4 <- mstmap(mapBC3, bychr = FALSE, trace = FALSE, dist.fun = "kosambi",
                 p.value = 1e-12)
pg <- profileGen(mapBC4, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
                 id = "Genotype", xo.lambda = 14, layout = c(1, 3), lty = 2,
                 cex = 0.7)
```

These notes concern aspects of the MSTmap algorithm, and of the arguments
supplied to the construction functions, that emerge only from detailed use.
Each has the same practical consequence: apparently inflated genetic distances.

## Inflation of genetic distances

There are circumstances in which the algorithm produces inflated genetic
distances. This is most pronounced when the map being constructed contains
genotypes with many missing values, since the constructed map will then contain
runs of missing values for those genotypes.

```{r dist1, fig.cap = "Missing allele scores for the constructed map mapBC4.", fig.alt = "Missing allele scores for the constructed map mapBC4."}
plotMissing(mapBC4)
```

The dark lines show the phenomenon in the initial map `mapBC4`. Seven of these
genotypes were eventually removed for excessive recombination across the genome,
but their missing value structure contributed substantially to the inflated
lengths of the linkage groups in its own right.

The mechanism is most easily shown by reconstructing the map with
`return.imputed = TRUE`, which adds the imputed marker probability matrix to the
returned object.

```{r dist3}
mapBC4i <- mstmap(mapBC3, bychr = FALSE, trace = FALSE, dist.fun = "kosambi",
                  p.value = 1e-12, return.imputed = TRUE)
mapBC4i$geno[[1]]$map[1:14]
mapBC4i$imputed.geno[[1]]$map[1:5]
```

Consider the markers on L.1. The first 14 markers contain the first two
co-locating sets and are calculated to lie 0.76 cM apart. The imputed map is a
reduced form containing all unique markers together with one member of each
co-locating set; `mark246` and `mark391` are the members chosen for the first
two sets. The number of observed allele pairs between the two markers, and the
number of observed recombinations across those pairs, may be counted directly.

```{r dist4}
len <- apply(mapBC4$geno[[1]]$data[, c(1, 5)], 1, function(el)
             length(el[!is.na(el)]))
length(len[len > 1])
bca <- apply(mapBC4i$geno[[1]]$data[, c(1, 5)], 1, function(el) {
    el <- el[!is.na(el)]
    sum(abs(diff(el)))
})
bca[bca > 0]
```

A single genuine recombination among 281 observed pairs implies a distance of
0.35 cM, well short of the 0.76 cM reported. The shortfall is explained by the
imputed marker probabilities for the first five markers of the seven genotypes
carrying excessive missing values.

```{r dist5}
mapBC4i$imputed.geno[[1]]$data[pg$xo.lambda, 1:5]
```

Genotype BC061 has an extended run of missing values across the first four
co-located sets. Under the distance formula in the M-step of the EM algorithm,
given as equation (5) in
[How the MSTmap algorithm works](mstmap-algorithm.html), the estimated
probabilities for its two missing scores in the first two markers contribute a
further half recombination, or approximately 0.15 cM, to the distance between
them. BC061 contributes similarly between the second and third sets and between
the third and fourth, because the imputed probabilities remain close to 0.5
throughout the run. Where several genotypes carry such runs, these small
contributions accumulate quickly and the linkage groups appear inflated.

Two remedies are available. The first is to remove the offending genotypes,
irrespective of their value elsewhere in the map. The second, less obvious, is
to re-estimate the distances with
[`quickEst()`](../reference/quickEst.html); because that function imputes
missing values with actual allele calls, it avoids the uncertain probabilities
from which MSTmap forms its distances.

```{r dist6}
mapBC4e <- quickEst(mapBC4)
chrlen(mapBC4)
chrlen(mapBC4e)
```

The reduction is substantial: in excess of 80 cM for L.1, and over 60 cM for
both L.2 and L.4.

> Where a map has been constructed from genotypes carrying many missing values,
> the missing value structure of the resulting object should be examined. If
> runs of missing values are present the linkage groups are most likely
> inflated, and the distances should be checked with `quickEst()`.

## The argument `mvest.bc`

Inflation of this kind can be made considerably worse by injudicious use of
`mvest.bc`.

The reason lies in how co-segregating markers are determined. In the initial
stages of the algorithm two markers are held to be co-located if their pairwise
distance is zero, that distance being computed only from allele pairs observed
in both markers. A representative is chosen from each co-locating set, the
remainder set aside with a record of their link to it, and the map constructed
from the representatives alone. This is the behaviour under
`mvest.bc = FALSE`.

Under `mvest.bc = TRUE`, missing allele calls are instead assigned a probability
of 0.5 of being an A allele *before* clustering takes place. This has the
evident advantage of involving the complete set of genotypes in the pairwise
calculations. It has a less evident consequence: for any two markers with some
proportion of unobserved allele pairs, the pairwise distance becomes non-zero,
and the markers are therefore not judged co-located. A larger set of markers
consequently enters construction, with the original missing calls imputed
repeatedly as part of marker ordering. Where missing values are concentrated in
a subset of genotypes, the cumulative distances between adjacent markers
increase quickly.

```{r mvest1}
mapBC4a <- mstmap(mapBC3, bychr = FALSE, trace = FALSE, dist.fun = "kosambi",
                  p.value = 1e-12, mvest.bc = TRUE)
nmar(mapBC4)
```

Clustering yields an identical number of linkage groups containing the same
markers. The number of uniquely located markers within each group, however,
increases considerably, and with it the length of each group.

```{r mvest2}
sapply(mapBC4a$geno, function(el) length(unique(round(el$map, 4)))) -
  sapply(mapBC4$geno, function(el) length(unique(round(el$map, 4))))
chrlen(mapBC4a)
```

The remedy is again `quickEst()`.

```{r mvest3}
mapBC4b <- quickEst(mapBC4a)
chrlen(mapBC4b)
```

## The argument `detectBadData`

The preceding sections concern inflation arising from missing values. Where
missing values are not the cause and distances remain inflated, the explanation
is most likely an increased recombination rate among some or all genotypes,
which is readily checked with `statGen()` or `profileGen()` once an initial map
exists.

An increased rate confined to a few genotypes is dealt with by removing them.
A general inflation across many or all genotypes is more awkward, and suggests
genotyping errors introduced during the physical process of obtaining the allele
calls.

Setting `detectBadData = TRUE` instructs MSTmap to identify suspiciously called
alleles, by the criterion given as equation (6) in
[How the MSTmap algorithm works](mstmap-algorithm.html). Suspicious calls are
set to missing and imputed by the EM algorithm, which reduces the number of
recombinations between adjacent markers and may reduce both those distances and
the overall linkage group lengths substantially.

The behaviour is illustrated by introducing simulated genotyping errors into the
final map `mapBC`, switching individual allele calls at random within each
linkage group.

```{r dbd1}
mapBCd <- mapBC
mapBCd$geno <- lapply(mapBCd$geno, function(el) {
    ns <- sample(1:ncol(el$data), ncol(el$data) / 2, replace = TRUE)
    ns <- cbind(sample(1:nrow(el$data), ncol(el$data) / 2, replace = TRUE), ns)
    el$data[ns] <- abs(1 - el$data[ns])
    el$data[el$data == 0] <- 2
    el
})
mapBCd <- quickEst(mapBCd)
chrlen(mapBCd)
```

`quickEst()` does not disregard the introduced errors, and the linkage groups
are correspondingly inflated.

```{r dbd2}
mapBCda <- mstmap(mapBCd, bychr = TRUE, trace = FALSE, dist.fun = "kosambi",
                  p.value = 1e-12, detectBadData = TRUE)
chrlen(mapBCda)
```

Detecting and disregarding the errors during marker ordering returns the linkage
groups to lengths comparable with those of the final map `mapBC`.

> `detectBadData` alters the data on which distances are based, and will reduce
> linkage group lengths whether or not genotyping errors are truly present. It
> should be applied deliberately, once other explanations for inflation have
> been excluded, rather than as a routine setting.

## Further reading

- [How the MSTmap algorithm works](mstmap-algorithm.html) — the imputation and
  error detection procedures these arguments control
- [Manipulating linkage maps](manipulating-maps.html) — `quickEst()` in detail
- [Worked example I](worked-example-construction.html) — the map these notes
  refer to

## References
