---
title: "Constructing a linkage map"
description: >
  The two construction functions, for a data frame of marker scores and for an R/qtl cross object, and the MSTmap parameters they share.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```


This chapter explores the R/ASMap functions in greater depth and shows how they can be used to efficiently explore, manipulate and construct genetic linkage maps. It will also showcase the graphical tools that will allow efficient diagnosis of linkage map problems post construction.

The package contains multiple data sets listed as follows

**mapDH**

:   : A constructed genetic linkage map for a Doubled Haploid population in the form of an R/qtl object. The genetic linkage map contains a total of 599 markers spanning 23 linkage groups genotyped across 218 individuals. The linkage map contains a small set of co-located markers and a small set of markers with excessive segregation distortion

**mapDHf**

:   : An unconstructed version of `mapDH` in the form a data frame. The data frame has dimensions 599 $\times$ 218 and the rows (markers) have been randomized.

**mapBCu**

:   : An unconstructed set of markers for a backcross population in the form of an R/qtl object. The marker set contains a total of 3023 markers genotyped on 326 individuals. This marker set can be used in conjunction with the detailed process presented in chapter to construct a genetic linkage map.

**mapBC**

:   : A constructed linkage map of `mapBCu` in the form of an R/qtl object. The linkage map contains a total of 3019 markers genotyped on 300 individuals.

**mapF2**

:   : A simulated linkage map for a self-pollinated F2 population consisting of 700 markers spanning 7 linkage groups genotyped across 250 individuals.

Each of these data sets is accessible using commands similar to

```{r data0}
data(mapDHf, package = "ASMap")
data(mapDH, package = "ASMap")
data(mapBCu, package = "ASMap")
```

## Map construction functions

The R/ASMap package contains two linkage map construction functions that allow users to fully utilize the MSTmap parameters listed at <http://alumni.cs.ucr.edu/~yonghui/mstmap.html>. Some additional information on aspects of the MSTmap algorithm and the appropriate use of the arguments `mvest.bc` and `detectBadData` is given in Chapter .

### `mstmap.data.frame()`

The first of these functions allows users to input a data frame of genetic markers ready for construction. For a more detailed explanation of the arguments users should consult the help documentation found by typing `?mstmap.data.frame` in R.

```r
mstmap.data.frame(object, pop.type = "DH", dist.fun = "kosambi",
      objective.fun = "COUNT", p.value = 1e-06, noMap.dist = 15,
      noMap.size = 0, miss.thresh = 1, mvest.bc = FALSE, detectBadData = FALSE,
      as.cross = TRUE, return.imputed = TRUE, trace = FALSE, ...)
```

The explicit form of the data frame `object` is borne from the syntax of the marker file required for using the stand alone MSTmap software. It must have markers in rows and genotypes in columns. Marker names are required to be in the `rownames` component of the object with genotype names residing in the `names`. Spaces in any of the marker or genotype names should be avoided but will be replaced with a “-” if found. Each of the columns of the data frame must be of class `"character"` (not factors). If converting from a matrix, this can easily be achieved by using the `stringAsFactors = FALSE` argument for any `data.frame` method.

The available populations that can be passed to `pop.type` are `"BC"` Backcross, `"DH"` Doubled Haploid, `"ARIL"` Advanced Recombinant Inbred and `"RILn"` Recombinant Inbred with n levels of selfing. The allelic content of the markers in the `object` must be explicitly adhered to. For `pop.type` `"BC"`, `"DH"` or `"ARIL"` the two allele types should be represented as (`"A"` or `"a"`) and (`"B"` or `"b"`). Thus for `pop.type = "ARIL"` it is assumed the minimal number of heterozygotes have been set to missing. For non-advanced RIL populations (`pop.type = "RILn"`) phase unknown heterozygotes should be represented as `"X"`. For all populations, missing marker scores should be represented as (`"U"` or `"-"`).

Users need to be aware that the `p.value` argument plays an important role in determining the clustering of markers to distinct linkage groups. Section Clustering shows the separation of marker groups is highly dependent on the the number of individuals in the population. For this reason, some trial and error may be required to determine an appropriate `p.value` for the linkage map being constructed.

Although this function contains arguments that utilize the complete set of available MSTmap parameters it is less flexible than its sister function `mstmap.cross()` (see section textttmstmap.cross()) that uses the flexible structure of an R/qtl `"cross"` object. For this reason, it is recommended that users set `as.cross = TRUE` to ensure the constructed object is returned as a R/qtl cross object with an appropriate class structure. For population types `"BC"` and `"DH"` the class of the constructed object is given `"bc"` and `"dh"` respectively. For `"RILn"` the **qtl** package conversion function `convert2bcsft` is used to ensure the class of the object is assigned `"bcsft"` with arguments `F.gen = n` and `BC.gen = 0`. For `"ARIL"` populations the constructed object is given the class `"riself"`. The correct assignation of these classes ensures the objects can be used synergistically with the suite of functions available in the R/qtl package as well as other functions in the R/ASMap package.

The R/ASMap package contains an unconstructed Doubled Haploid marker set `mapDHf` with 599 markers genotyped across 218 individuals. The marker set is formatted correctly for input into the `mstmap.data.frame()` function.

```{r datadf}
testd <- mstmap(mapDHf, dist.fun = "kosambi", trace = TRUE, as.cross = TRUE)
nmar(testd)
chrlen(testd)
```

As `as.cross = TRUE` the usual functions available in the R/qtl package are available for use on the returned object.

### `mstmap.cross()`

The second linkage map construction function allows users to input an unconstructed or constructed linkage map in the form of an R/qtl cross object. See `?mstmap.cross` for a more detailed description.

```r
mstmap.cross(object, chr, id = "Genotype", bychr = TRUE,
       suffix = "numeric", anchor = FALSE, dist.fun = "kosambi",
       objective.fun = "COUNT", p.value = 1e-06, noMap.dist = 15,
       noMap.size = 0, miss.thresh = 1, mvest.bc = FALSE, detectBadData =
       FALSE, return.imputed = FALSE, trace = FALSE, ...)
```

The cross `object` needs to inherit from one of the allowable classes available in the R/qtl package, namely `"bc","dh","riself","bcsft"` where `"bc"` is a Backcross `"dh"` is Doubled Haploid, `"riself"` is an advanced Recombinant Inbred and `"bcsft"` is a Backcross/Self.

It is important to understand how these classes are encoded into the object for the specific populations. If `read.cross()` is used to read in any bi-parental populations it will be given the class `"bc"`. Doubled Haploid populations can be changed to `"dh"` just by changing the class. For the purpose of linkage map construction, both classes `"bc"` and `"dh"` will produce equivalent results. For non-advanced Recombinant Inbred populations markers are required to be fully informative (i.e. contain 3 distinct allele types such as AA, BB for parental homozygotes and AB for phase unknown heterozygotes) and the use of `read.cross()` will result in the cross object being given a class `"f2"`. The level of selfing is required to be encoded into the object by applying one of the two conversion functions available in the R/qtl package. For a population that has been generated by selfing $n$ times, the conversion function `convertbcsft` can be used by setting the arguments `F.gen = n` and `BC.gen = 0`. This will attach a class `"bcsft"` to the object. Populations that are genuine advanced RILs can be converted using the `convert2riself` function. This function will replace any remaining heterozygosity, if it exists, with missing values and attach the class `"riself"` to the object.

Similar to the `mstmap.data.frame()` function, users need to be aware that the `p.value` argument is highly dependent on the number of individuals in the population and may require some trial and error to ascertain an appropriate value. After construction the cross object is returned with an identical class structure as the inputted object. All R/qtl and R/ASMap package functions can be used synergistically with this object.

#### Examples

The constructed linkage map `mapDH` available in the R/ASMap package will be used to showcase the flexibility of this function. Before attempting re-construction, some preliminary output of `mapDH` is presented.

```{r data}
nmar(mapDH)
pull.map(mapDH)[[4]]
```

The output shows that there are 23 groups that have been appropriately been assigned linkage group or chromosome names. The markers within each linkage group have been named according to the order of the markers.

**Example 1: Completely construct or reconstruct a linkage map.**

To completely re-construct this map set the argument `bychr = FALSE`. This will bulk the genetic data from all linkage groups, re-cluster the markers into groups and then optimally order the markers within each linkage group. This linkage map is small so these two tasks happen almost instantaneously.

```{r mst1}
mapDHa <- mstmap(mapDH, bychr = FALSE, dist.fun = "kosambi", trace = TRUE)
nmar(mapDHa)
pull.map(mapDHa)[[4]]
```

The reconstructed map contains 24 linkage groups with the extra linkage group coming from a minor split in 4A. As the linkage map is constructed from scratch, it assumed that former linkage group names are no longer required. A standard “L.” prefix is provided for the new linkage group names. This example also indicates that MSTmap by default does not respect the inputted marker order.

**Example 2: Optimally order markers within linkage groups of an established map.**

It may be necessary to only perform marker ordering within already established linkage groups. This can be achieved by setting `bychr = TRUE`. In some cases it also be preferable to ensure the marker orders of the linkage groups are respected and this can be achieved by setting `anchor = TRUE.`

```{r mst2}
mapDHb <- mstmap(mapDH, bychr = TRUE, dist.fun = "kosambi", anchor = TRUE, trace = TRUE)
nmar(mapDHb)
```

This map is identical to `mapDH` with the exception that chromosome 4A has been split into two linkage groups. As `bychr = TRUE` the function understands the origin of the linkage group was 4A and consequently uses it as a prefix in the naming of the two new linkage groups.

**Example 3: Optimally order markers within linkage groups of an established map without breaking linkage groups.**

The splitting of the linkage groups in the last example only occurred due to choice of default `p.value = 1e-06` set in the function. A slight change to this `p.value` will ensure that 4A remains linked during the algorithm.

```{r mst3}
mapDHc <- mstmap(mapDH, bychr = TRUE, dist.fun = "kosambi", anchor = TRUE, trace = TRUE, p.value = 1e-04)
nmar(mapDHc)
```

An identical result can be achieved by setting `p.value = 2` or any number greater than 1. Doing this instructs the MSTmap algorithm to not split linkage groups regardless of how weak the linkages are between markers within any group. Users should be aware that if this latter method is used then linkage groups that contain groups of markers separated by a substantial distance (i.e. very weak linkages) may suffer from local orientation problems.

**Example 4: Reconstruct map within predefined linkage groups of an established map.**

There may only be a need to reconstruct a predefined set of linkage groups. By setting the `chr` argument and `bychr = FALSE` users can determine which linkage groups require the complete reconstruction using MSTmap.

```{r mst4}
mapDHd <- mstmap(mapDH, chr = names(mapDH$geno)[1:3], bychr = FALSE, dist.fun = "kosambi", trace = TRUE, p.value = 1e-04)
nmar(mapDHd)
```

Again, the algorithm assumes that the original linkage group names are no longer necessary and defines new ones. An obvious extension of this example is to set `bychr = TRUE` and then the algorithm will order markers within the predefined linkage groups stipulated by `chr`. Similar to the previous example, an appropriate choice of `p.value` will ensure that linkage groups remain unbroken.

