---
title: "Constructing a linkage map"
description: >
  The two construction functions, for a data frame of marker scores and for an
  R/qtl cross object, and the arguments that govern their behaviour.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBCu)
```

ASMap provides two construction functions, both methods of the `mstmap()`
generic. They differ only in the object they accept: a data frame of marker
scores, or an R/qtl `cross` object. Both expose the full set of MSTmap
parameters documented at <http://alumni.cs.ucr.edu/~yonghui/mstmap.html>.

The algorithm underlying them is described in
[How the MSTmap algorithm works](mstmap-algorithm.html), and the practical
behaviour of the `mvest.bc` and `detectBadData` arguments in
[Notes on the MSTmap algorithm](algorithm-notes.html).

## Example data

The package supplies five data sets, described in full on their reference
pages. The datasets are not lazy-loaded and require an explicit `data()` call.

| Data set | Form | Individuals | Markers | Description |
|---|---|---|---|---|
| [`mapDH`](../reference/mapDH.html) | cross | 218 | 599 | Constructed doubled haploid map, 23 linkage groups, containing a few co-located and distorted markers |
| [`mapDHf`](../reference/mapDHf.html) | data frame | 218 | 599 | Unconstructed form of `mapDH`, with the marker rows randomised |
| [`mapBCu`](../reference/mapBCu.html) | cross | 326 | 3023 | Unconstructed backcross marker set |
| [`mapBC`](../reference/mapBC.html) | cross | 300 | 3019 | Constructed form of `mapBCu` |
| [`mapF2`](../reference/mapF2.html) | cross | 250 | 700 | Simulated selfed F2 map, 7 linkage groups |

```{r data0}
data(mapDHf, package = "ASMap")
data(mapDH, package = "ASMap")
data(mapBCu, package = "ASMap")
```

## Construction from a data frame

[`mstmap.data.frame()`](../reference/mstmap.data.frame.html) accepts a data
frame of marker scores. Its required layout derives from the marker file used by
the standalone MSTmap software: **markers in rows and genotypes in columns**,
with marker names in the `rownames` and genotype names in the `names`. Every
column must be of class `"character"` rather than a factor, which is most easily
ensured through the `stringsAsFactors = FALSE` argument of any `data.frame`
method. Spaces in marker or genotype names should be avoided; where found they
are replaced by a hyphen.

The allelic coding is strict, and depends on the population type given to
`pop.type`.

| `pop.type` | Population | Allele coding |
|---|---|---|
| `"BC"` | Backcross | `"A"` or `"a"`, and `"B"` or `"b"` |
| `"DH"` | Doubled haploid | `"A"` or `"a"`, and `"B"` or `"b"` |
| `"ARIL"` | Advanced recombinant inbred | `"A"` or `"a"`, and `"B"` or `"b"`; heterozygotes assumed already set to missing |
| `"RILn"` | Recombinant inbred, *n* levels of selfing | As above, with phase-unknown heterozygotes coded `"X"` |

Missing scores are denoted `"U"` or `"-"` for all population types.

`mapDHf` is an unconstructed doubled haploid marker set of 599 markers scored
on 218 individuals, formatted for input to this function.

```{r datadf}
testd <- mstmap(mapDHf, dist.fun = "kosambi", trace = FALSE, as.cross = TRUE)
nmar(testd)
chrlen(testd)
```

Setting `as.cross = TRUE` is recommended, since it returns an R/qtl cross object
of the appropriate class and thereby makes the facilities of both packages
available for subsequent work. The class assigned depends on the population
type.

| `pop.type` | Class assigned | Mechanism |
|---|---|---|
| `"BC"` | `"bc"` | Direct |
| `"DH"` | `"dh"` | Direct |
| `"ARIL"` | `"riself"` | Direct |
| `"RILn"` | `"bcsft"` | Via `qtl::convert2bcsft()` with `F.gen = n`, `BC.gen = 0` |

With `as.cross = FALSE` the map is returned instead as a data frame carrying
additional columns for the linkage group, marker position and genetic distance.

## Construction from a cross object

[`mstmap.cross()`](../reference/mstmap.cross.html) accepts an unconstructed or
constructed map held in an R/qtl cross object, and is the more flexible of the
two. The object must inherit one of the classes `"bc"`, `"dh"`, `"riself"` or
`"bcsft"`, a restriction that guards against attempting construction for the
more complex population types R/qtl supports.

How these classes arise matters in practice. `read.cross()` assigns the class
`"bc"` to any bi-parental population. Doubled haploid populations may be
assigned `"dh"` simply by changing the class; for the purpose of construction
the two are equivalent.

Non-advanced recombinant inbred populations require fully informative markers,
that is three distinct allele types, and `read.cross()` will assign them the
class `"f2"`. The level of selfing must then be encoded by conversion:

| Population | Conversion |
|---|---|
| Selfed $n$ times | `qtl::convert2bcsft(F.gen = n, BC.gen = 0)`, giving class `"bcsft"` |
| Genuine advanced RIL | `qtl::convert2riself()`, giving class `"riself"` and replacing any residual heterozygosity with missing values |

The object is returned with the same class structure it was given, so that all
functions of both packages remain applicable.

> For both functions, the `p.value` argument governs the separation of markers
> into linkage groups and is strongly dependent on population size. Some
> experimentation is generally required. See
> [Clustering markers into linkage groups](mstmap-algorithm.html) for the
> relationship, which may be examined directly with `pValue()`.

## Modes of reconstruction

The constructed map `mapDH` illustrates the flexibility of `mstmap.cross()`.
Its 23 linkage groups are named, and the markers within each are named according
to their order.

```{r data}
nmar(mapDH)
pull.map(mapDH)[[4]]
```

The four modes below are selected through `bychr`, `chr` and `anchor`.

| Requirement | `bychr` | `chr` | Effect on linkage groups |
|---|---|---|---|
| Rebuild the entire map | `FALSE` | unset | Re-clustered and renamed |
| Re-order within existing groups | `TRUE` | unset | Retained, possibly split |
| Rebuild nominated groups only | `FALSE` | set | Nominated groups re-clustered |
| Re-order nominated groups only | `TRUE` | set | Nominated groups retained |

### Complete reconstruction

Setting `bychr = FALSE` combines the marker data from all linkage groups,
re-clusters the markers and orders them within each resulting group.

```{r mst1}
mapDHa <- mstmap(mapDH, bychr = FALSE, dist.fun = "kosambi", trace = FALSE)
nmar(mapDHa)
pull.map(mapDHa)[[4]]
```

The reconstructed map contains 24 linkage groups, the additional group arising
from a minor split in 4A. Because the map is built from first principles the
former linkage group names are discarded and a standard `"L."` prefix applied.
This example also shows that MSTmap does not, by default, respect the marker
order it was given.

### Ordering within established groups

Where only the order within existing linkage groups is at issue, set
`bychr = TRUE`. Setting `anchor = TRUE` additionally causes the inputted marker
orders to be respected.

```{r mst2}
mapDHb <- mstmap(mapDH, bychr = TRUE, dist.fun = "kosambi", anchor = TRUE, trace = FALSE)
nmar(mapDHb)
```

The result is identical to `mapDH` except that 4A has been split in two. As
`bychr = TRUE`, the function recognises the origin of the group and uses 4A as a
prefix in naming the two new groups.

### Ordering without splitting

The split above arose solely from the default `p.value = 1e-06`. Relaxing it
keeps 4A intact.

```{r mst3}
mapDHc <- mstmap(mapDH, bychr = TRUE, dist.fun = "kosambi", anchor = TRUE,
                 trace = FALSE, p.value = 1e-04)
nmar(mapDHc)
```

The same result follows from `p.value = 2`, or any value greater than unity,
which instructs the algorithm not to split linkage groups however weak the
linkages within them.

> Suppressing splitting entirely carries a cost. A linkage group containing sets
> of markers separated by a substantial distance, and therefore weakly linked,
> may suffer local orientation problems.

### Reconstruction of nominated groups

Supplying `chr` together with `bychr = FALSE` restricts complete reconstruction
to the nominated linkage groups.

```{r mst4}
mapDHd <- mstmap(mapDH, chr = names(mapDH$geno)[1:3], bychr = FALSE,
                 dist.fun = "kosambi", trace = FALSE, p.value = 1e-04)
nmar(mapDHd)
```

Here too the original group names are discarded and new ones defined. Setting
`bychr = TRUE` instead orders markers within the groups nominated by `chr`, and
an appropriate `p.value` will keep those groups unbroken.

## Further reading

- [How the MSTmap algorithm works](mstmap-algorithm.html) — the basis of
  `objective.fun`, `p.value`, `mvest.bc` and `detectBadData`
- [Pulling and pushing markers](pulling-and-pushing.html) — curating the marker
  set before construction
- [Worked example I](worked-example-construction.html) — construction applied to
  an unconstructed barley backcross
- [`mstmap.cross()`](../reference/mstmap.cross.html),
  [`mstmap.data.frame()`](../reference/mstmap.data.frame.html) — full argument
  documentation

## References
