---
title: "Diagnosing genotypes and markers"
description: >
  Numerical and graphical profiling of individual genotypes, markers and intervals, and the identification of genetic clones.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```

## Genotype and marker/interval profiling

To provide a complete system for efficient linkage map construction and diagnosis the R/ASMap package contains functions that calculate linkage map statistics across the markers for every genotype as well as across the markers/intervals of the genome. These statistics can then be profiled across simultaneous R/lattice panels for quick diagnosis of linkage map attributes.

```r
statGen(cross, chr, bychr = TRUE, stat.type = c("xo", "dxo", "miss"), id = "Genotype")
profileGen(cross, chr, bychr = TRUE, stat.type = c("xo", "dxo", "miss"), id = "Genotype", xo.lambda = NULL, ...)
statMark(cross, chr, stat.type = c("marker", "interval"), map.function = "kosambi")
profileMark(cross, chr, stat.type = "marker", use.dist = TRUE, map.function = "kosambi", crit.val = NULL, display.markers = FALSE, mark.line = FALSE, ...)
```

The `statGen()` and `profileGen()` are functions for calculation and profiling of the statistics for the genotypes across the chosen marker set defined by `chr`. Specifically, `profileGen()` actually calls `statGen()` to obtain the statistics for profiling and also returns the statistics invisibly after plotting. Setting `bychr = TRUE` will also ensure that the genotype profiles are plotted individually for each linkage groups given `chr`. The current statistics that can be calculated for each genotype include

::: list
$\bullet$

`"xo"` : number of crossovers

`"dxo"` : number of double crossovers

`"miss"` : number of missing values
:::

The two statistics `"xo"` and `"dxo"` are obviously only useful for constructed linkage maps. However, from my experience, they represent the most vital two statistics for determining a linkage maps quality. Inflated crossover or double crossover rates of any genotypes indicate a lack of adherence to Mendelian genetics. For example, it can be quickly calculated for a Doubled Haploid wheat population of size 100, one recombination is approximately 1cM. Consequently, under Mendelian genetics, a chromosome of length 200 cM would expect to have 200 random crossovers with each genotype having an expected recombination rate of 2. Wheat is a hexaploid, so for excellent coverage over the whole genome the expected recombination rate of any genotype is $\sim$`<!-- -->`{=html}42. This number will obviously be reduced if the coverage of the genome is incomplete. Significant recombination rates can be checked by manually inputting a median recombination rate in the argument `xo.lambda` of `profileGen()`. As an example, the genotype profiles of the `mapDH` can be displayed using the command below and the resultant plot is given in the figure below.

```{r prof2, fig.width = 10.5, fig.height = 5.6, fig.cap = "Genotype profiles of missing values, double recombinations and recombinations for `mapDH`."}
profileGen(mapDH, bychr = FALSE, stat.type = c("xo", "dxo", "miss"), id = "Genotype", xo.lambda = 25, layout = c(1,3), lty = 2)
```

The plot neatly displays, the number of missing values, double crossovers and crossovers for each genotype in the order represented in `mapDH`. The plot was aesthetically enhanced by including graphical arguments `layout = c(1,3)` and `lty = 2` which are passed to the high level lattice function `xyplot()`. Plotting these statistics simultaneously allows the users to quickly recognize genotypes that may be problematic and worth further investigation. For example, in the figure below the line DH218 was identified as having an inflated recombination rate and removing it may improve the quality of the linkage map.

The `statMark()` and `profileMark()` functions are commands for the calculation and profiling of statistics associated with the markers or intervals of the linkage map. `profileMark()` calls `statMark()` and graphically profiles user given statistics and will also return them invisibly after plotting. The current marker statistics that can be profiled are

::: list
$\bullet$

`"seg.dist"`: -log10 p-value from a test of segregation distortion

`"miss"`: proportion of missing values

`"prop"`: allele proportions

`"dxo"`: number of double crossovers
:::

The current interval statistics that can be profiled are

::: list
$\bullet$

`"erf"`: estimated recombination fractions

`"lod"`: LOD score for the test of no linkage

`"dist"`: interval map distance

`"mrf"`: map recombination fraction

`"recomb"`: number of recombinations.
:::

The function allows any or all of the statistics to be plotted simultaneously on a multi-panel lattice display. This includes combinations of marker and interval statistics. There is a `chr` argument to subset the linkage map to user defined linkage groups. If `crit.val = "bonf"` then markers that have significant segregation distortion greater than the family wide alpha level of 0.05/no.of.markers will be annotated in marker panels. Similarly, intervals that have a significantly weak linkage from a test of the recombination fraction of $r = 0.5$ will also be annotated in the interval panels. Similar to `profileGen()`, graphical arguments can be used in `profileMark()` and are passed to the high level lattice function `xyplot()`. A plot of the segregation distortion, double crossovers, estimated recombination fractions and LOD scores for the whole genome of `mapDH` is given in the figure below and created with the command

```{r prof3, fig.width = 10.5, fig.height = 5.6, fig.cap = "Marker and interval profiles of segregation distortion, double crossovers, estimated recombination fractions and LOD scores for `mapDH`.", warning = FALSE}
profileMark(mapDH, stat.type = c("seg.dist", "dxo", "erf", "lod"), id = "Genotype", layout = c(1,4), type = "l")
```

All linkage groups are highlighted in a different colours to ensure they can be identified clearly. The lattice panels ensure that marker and interval statistics are seamlessly plotted together so problematic regions or markers can be identified efficiently. In this command, the `layout = c(1,4)` and `type = "l"` arguments are passed directly to the high level `xyplot()` lattice function to ensure a more aesthetically pleasing graphic.

### Genetic clones

In my experience, the assumptions of how the individuals of the population are genetically related is rarely checked throughout the construction process. Too often unconstructed or constructed linkage maps contain individuals that are far too closely related beyond the simple assumptions of the population. For example, in a DH population, the assumption of independence would indicate that any two individuals will, by chance, share half of their alleles. Any pairs of individuals that significantly breach this assumption should be deemed suspicious and queried.

The R/ASMap package contains a function for the detection and reporting of the relatedness between individuals as well as a function for forming consensus genotypes if genuine clones are found.

```r
genClones(object, chr, tol = 0.9, id = "Genotype")
fixClones(object, gc, id = "Genotype", consensus = TRUE)
```

The `genClones()` function uses the power of `comparegeno()` from the R/qtl package to perform the relatedness calculations. It then provides a numerical breakdown of the relatedness between pairs of individuals that share a proportion of alleles greater than `tol`. This breakdown also includes the clonal group the pairs of individuals belong to. The complete set of statistics allows users to make an informed decision about the connectedness of the pairs of individuals. For example, using the constructed map `mapDH` the list of clones can be found using

```{r clones02}
gc <- genClones(mapDH, tol = 0.9)
gc$cgd
```

The reported information shows five pairs of clones that are, with the exception of missing values, identical.

If clones are found then `fixClones()` can be used to form consensus genotypes for each of the clonal groups. By default it will intelligently collapse the allelic information of the clones within each group (see the table below) to obtain a single consensus genotype. Setting `consensus = FALSE` will choose a genotype with the smallest proportion of missing values as the representative genotype for the clonal group. In both cases genotypes names are given an elongated name containing all genotype names of the clonal group separated by an underscore.

```{r clones03}
mapDHg <- fixClones(mapDH, gc$cgd, consensus = TRUE)
levels(mapDHg$pheno[[1]])[grep("_", levels(mapDHg$pheno[[1]]))]
```

  Genotype    M1   M2   M3   M4   M5   M6   M7   M8
  ---------- ---- ---- ---- ---- ---- ---- ---- ----
  G1          AA   BB   AA   BB   AA   BB   NA   AA
  G2          AA   BB   NA   NA   NA   NA   NA   BB
  G3          AA   BB   AA   BB   NA   NA   NA   AA
  Cons.       AA   BB   AA   BB   AA   BB   NA   NA

  : Consensus genotype outcomes for 3 clones across 8 markers in a DH population.

