---
title: "Diagnosing genotypes and markers"
description: >
  Numerical and graphical profiling of individual genotypes, of markers and of
  intervals, and the detection of genetic clones.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

ASMap provides functions that calculate linkage map statistics both across the
markers for every genotype and across the markers and intervals of the genome.
These may then be profiled on simultaneous lattice panels, which permits the
attributes of a map to be diagnosed rapidly.

In each case a calculation function is paired with a profiling function. The
profiling function calls its numerical counterpart, plots the result, and
returns the statistics invisibly.

| Calculation | Profiling | Applies to |
|---|---|---|
| [`statGen()`](../reference/statGen.html) | [`profileGen()`](../reference/profileGen.html) | Individual genotypes |
| [`statMark()`](../reference/statMark.html) | [`profileMark()`](../reference/profileMark.html) | Markers and intervals |

## Genotype statistics

Three statistics may be calculated for each genotype across the marker set
nominated by `chr`.

| `stat.type` | Statistic |
|---|---|
| `"xo"` | Number of crossovers |
| `"dxo"` | Number of double crossovers |
| `"miss"` | Number of missing values |

Setting `bychr = TRUE` profiles the genotypes separately for each linkage group.

The crossover statistics apply only to constructed maps, but they are the two
most informative indicators of map quality. An inflated crossover or double
crossover rate for a genotype indicates a departure from the assumptions of
Mendelian genetics.

The expected rate can be derived for a given population. In a doubled haploid
wheat population of size 100, one recombination corresponds to approximately
1 cM; a chromosome of length 200 cM would therefore be expected to exhibit 200
random crossovers, giving each genotype an expected recombination rate of 2.
Wheat being hexaploid, complete coverage of the genome implies an expected rate
of approximately 42 for any genotype. Incomplete coverage reduces this figure.

Genotypes departing significantly from an expected rate are identified by
supplying that rate to the `xo.lambda` argument of `profileGen()`.

```{r prof2, fig.width = 10.5, fig.height = 5.6, fig.cap = "Genotype profiles of missing values, double recombinations and recombinations for `mapDH`.", fig.alt = "Genotype profiles of missing values, double recombinations and recombinations for mapDH."}
profileGen(mapDH, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
           id = "Genotype", xo.lambda = 25, layout = c(1, 3), lty = 2)
```

The three statistics are displayed for each genotype in the order in which they
appear in `mapDH`. Plotting them together allows problematic genotypes to be
recognised at once: the line DH218 is identified here as having an inflated
recombination rate, and its removal would be expected to improve the quality of
the map. The `layout` and `lty` arguments are passed to the underlying lattice
function `xyplot()`.

> `profileGen()` returns its statistics invisibly, including a logical vector
> named `xo.lambda` marking the genotypes it has flagged. This pairs directly
> with [`subsetCross()`](../reference/subsetCross.html), which removes the
> offending individuals while keeping any pulled-marker elements consistent.

## Marker and interval statistics

`profileMark()` profiles statistics associated with the markers or with the
intervals between them. Any combination may be displayed simultaneously,
including combinations drawn from both groups.

| Marker `stat.type` | Statistic |
|---|---|
| `"seg.dist"` | $-\log_{10}$ p-value from a test of segregation distortion |
| `"miss"` | Proportion of missing values |
| `"prop"` | Allele proportions |
| `"dxo"` | Number of double crossovers |

| Interval `stat.type` | Statistic |
|---|---|
| `"erf"` | Estimated recombination fraction |
| `"lod"` | LOD score for the test of no linkage |
| `"dist"` | Interval map distance |
| `"mrf"` | Map recombination fraction |
| `"recomb"` | Number of recombinations |

Setting `crit.val = "bonf"` annotates markers whose segregation distortion is
significant at a family-wise level of 0.05 divided by the number of markers,
and correspondingly annotates intervals exhibiting significantly weak linkage
under a test of $r = 0.5$. The map may be subset to nominated linkage groups
through `chr`.

```{r prof3, fig.width = 10.5, fig.height = 5.6, fig.cap = "Marker and interval profiles of segregation distortion, double crossovers, estimated recombination fractions and LOD scores for `mapDH`.", fig.alt = "Marker and interval profiles of segregation distortion, double crossovers, estimated recombination fractions and LOD scores for mapDH.", warning = FALSE}
profileMark(mapDH, stat.type = c("seg.dist", "dxo", "erf", "lod"),
            id = "Genotype", layout = c(1, 4), type = "l")
```

Each linkage group is drawn in a distinct colour. Marker and interval statistics
are plotted together on the same panels, so that problematic regions or
individual markers can be identified efficiently.

## Genetic clones

The assumptions made about the genetic relatedness of individuals are rarely
checked during construction, yet unconstructed and constructed maps not
uncommonly contain individuals far more closely related than the design of the
population implies. In a doubled haploid population, independence implies that
any two individuals will share half of their alleles by chance. Pairs that
depart substantially from this expectation warrant investigation.

[`genClones()`](../reference/genClones.html) uses `comparegeno()` from R/qtl
[@br14] to perform the relatedness calculation, and reports a numerical
breakdown for those pairs sharing a proportion of alleles greater than `tol`,
together with the clonal group to which each pair belongs.

```{r clones02}
gc <- genClones(mapDH, tol = 0.9)
gc$cgd
```

Five pairs are reported here which are identical but for their missing values.

> The table should be examined before it is acted upon. Pairs sharing few
> co-scored markers constitute weak evidence of cloning, and are better removed
> from the table than passed to `fixClones()`. The worked example demonstrates
> this.

[`fixClones()`](../reference/fixClones.html) forms consensus genotypes for the
clonal groups. By default the allelic information within each group is collapsed
to a single consensus genotype, in the manner shown below. Setting
`consensus = FALSE` instead retains the genotype with the smallest proportion of
missing values as the representative of the group. Under either setting the
resulting genotype is named by concatenating the names of the group, separated
by underscores.

```{r clones03}
mapDHg <- fixClones(mapDH, gc$cgd, consensus = TRUE)
levels(mapDHg$pheno[[1]])[grep("_", levels(mapDHg$pheno[[1]]))]
```

Consensus genotype outcomes for three clones across eight markers in a doubled
haploid population:

| Genotype | M1 | M2 | M3 | M4 | M5 | M6 | M7 | M8 |
|---|---|---|---|---|---|---|---|---|
| G1 | AA | BB | AA | BB | AA | BB | NA | AA |
| G2 | AA | BB | NA | NA | NA | NA | NA | BB |
| G3 | AA | BB | AA | BB | NA | NA | NA | AA |
| **Cons.** | AA | BB | AA | BB | AA | BB | NA | NA |

Marker M8 illustrates the rule: where the clones disagree, the consensus is set
to missing rather than resolved arbitrarily.

## Further reading

- [Heat maps](heat-maps.html) — the graphical counterpart to these numerical
  checks
- [Pulling and pushing markers](pulling-and-pushing.html) — acting on markers
  identified as problematic
- [Worked example I](worked-example-construction.html) — these diagnostics
  applied in sequence to an unconstructed marker set

## References
