---
title: "Getting started with ASMap"
description: >
  Construct a linkage map from a set of marker scores, and an orientation to
  the remainder of the documentation.
bibliography: ../ref.bib
link-citations: true
nocite: |
  @tb17
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf)
```

This article constructs a linkage map from a set of unordered marker scores,
discusses the single argument that most influences the outcome, and indicates
how the resulting map may be assessed. It is intended as an entry point to the
remaining documentation.

Construction is performed by the MSTmap algorithm [@mst08], which clusters
markers into linkage groups and determines an optimal order within each group
in a single pass. This is in contrast to the two-stage procedures common
elsewhere, in which an initial non-exhaustive ordering is followed by an
exhaustive search within a sliding marker window.

## Installation

```r
install.packages("ASMap")

# development version
# install.packages("devtools")
devtools::install_github("DrJ001/ASMap")
```

## Construction from a set of marker scores

`mapDHf` is an unconstructed doubled haploid wheat marker set, supplied in the
layout the construction function requires: **markers in rows, genotypes in
columns**, with marker names held in the `rownames` and genotype names in the
`names`.

> The package datasets are not lazy-loaded, and therefore require an explicit
> `data()` call. Its omission results in `object 'mapDHf' not found`.

```{r look}
library(ASMap)
data(mapDHf)

dim(mapDHf)
mapDHf[1:5, 1:6]
```

A single call clusters the markers into linkage groups, orders the markers
within each group and estimates the genetic distances between them.

```{r build}
map <- mstmap(mapDHf, pop.type = "DH", dist.fun = "kosambi", p.value = 1e-06)

nchr(map)
round(chrlen(map), 1)
```

The value returned is an R/qtl `cross` object, so the facilities of the
[qtl](https://CRAN.R-project.org/package=qtl) package [@br14] are immediately
available for further examination of the map.

```{r qtl, fig.height = 5}
summary(map)
plotMap(map)
```

## Construction from an existing cross object

Where the markers already reside in a `cross` object, the same generic accepts
it directly. Setting `bychr = FALSE` combines all markers and reconstructs the
map in its entirety, disregarding the existing linkage groups.

```{r cross}
data(mapDH)
map2 <- mstmap(mapDH, bychr = FALSE, dist.fun = "kosambi", p.value = 1e-06)
nchr(map2)
```

Setting `bychr = TRUE` instead retains the existing groups and re-orders the
markers within them. Two combinations of arguments are used repeatedly
throughout this documentation:

```r
# cluster and order from scratch, disregarding existing linkage groups
mstmap(cross, bychr = FALSE, p.value = 1e-12)

# re-order within existing linkage groups, without splitting them
mstmap(cross, bychr = TRUE, p.value = 2)
```

A `p.value` greater than unity suppresses clustering entirely, which is the
mechanism underlying the second of these.

## The choice of `p.value`

The `p.value` argument determines the threshold at which markers are separated
into distinct linkage groups. Its appropriate value depends strongly upon the
size of the population, larger populations requiring a smaller value, and some
experimentation is generally necessary. The relationship may be examined
directly with `pValue()`, which permits the choice to be made on a considered
basis.

```{r pvalue, fig.height = 5.5}
pValue(dist = c(25, 30, 35, 40), pop.size = 100:500, map.function = "kosambi")
```

For a population of approximately 300 individuals, a `p.value` of `1e-12`
corresponds to a threshold of some 30 cM before markers are assigned to a
common linkage group.

## Assessment of the constructed map

`heatMap()` displays the pairwise estimated recombination fractions and the
pairwise LOD scores of linkage together, each with its own legend. Consistent
heat within a linkage group indicates strong linkage between neighbouring
markers, and blocks of shared heat between groups suggest groups that may
require merging.

```{r heat, fig.height = 8, warning = FALSE}
heatMap(mapDH, lmax = 50)
```

## Further reading

| Article | Subject |
|---|---|
| [How the MSTmap algorithm works](mstmap-algorithm.html) | The minimum spanning tree formulation, clustering and marker ordering |
| [Constructing a linkage map](constructing-a-map.html) | Both construction functions and the MSTmap parameters |
| [Pulling and pushing markers](pulling-and-pushing.html) | Setting problematic markers aside and reintroducing them |
| [Diagnosing genotypes and markers](diagnostics.html) | Profiling of individuals, markers and intervals; genetic clones |
| [Heat maps](heat-maps.html) | Interpretation and tuning of the recombination fraction and LOD display |
| [Manipulating linkage maps](manipulating-maps.html) | Breaking, merging, subsetting and combining maps |
| [Worked example I: construction](worked-example-construction.html) | A barley backcross, from raw markers to a constructed map |
| [Worked example II: refinement](worked-example-refinement.html) | Reintroduction of markers, merging of groups, fine mapping |
| [Notes on the MSTmap algorithm](algorithm-notes.html) | Distance calculations, `mvest.bc` and `detectBadData` |

The package should be cited in published work as @tb17; `citation("ASMap")`
gives the current entry.

## References
