---
title: "Getting started with ASMap"
description: >
  Build a linkage map from a set of marker scores in a few lines, then find
  your way around the rest of the documentation.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf)
```

This article takes a set of raw marker scores and builds a finished linkage map
from them, in about five minutes. Along the way it covers the one argument you
will have to think about, and how to check that the result is any good.

ASMap does this with the MSTmap algorithm, which clusters markers into linkage
groups and orders them within each group in a single pass — no separate
"rippling" step, and no overnight wait.

## Installation

```r
install.packages("ASMap")

# development version
# install.packages("devtools")
devtools::install_github("DrJ001/ASMap")
```

## Building a map from marker scores

`mapDHf` is an unconstructed doubled haploid wheat marker set, in the layout the
construction function expects: **markers in rows, genotypes in columns**, with
marker names in the `rownames` and genotype names in the `names`.

> The package datasets are not lazy-loaded, so every example needs an explicit
> `data()` call. Forgetting it gives `object 'mapDHf' not found`.

```{r look}
library(ASMap)
data(mapDHf)

dim(mapDHf)
mapDHf[1:5, 1:6]
```

One call clusters the markers into linkage groups, orders them within each
group and estimates the genetic distances.

```{r build}
map <- mstmap(mapDHf, pop.type = "DH", dist.fun = "kosambi", p.value = 1e-06)

nchr(map)
round(chrlen(map), 1)
```

The result is an R/qtl `cross` object, so the whole of
[qtl](https://CRAN.R-project.org/package=qtl) is available to you immediately.

```{r qtl, fig.height = 5}
summary(map)
plotMap(map)
```

## Starting from an existing cross object

If the markers are already in a `cross` object, the same generic accepts it.
`bychr = FALSE` bulks every marker and rebuilds the map from scratch, ignoring
the existing linkage groups.

```{r cross}
data(mapDH)
map2 <- mstmap(mapDH, bychr = FALSE, dist.fun = "kosambi", p.value = 1e-06)
nchr(map2)
```

Setting `bychr = TRUE` instead keeps the existing groups and only re-orders the
markers within them. Two combinations are worth committing to memory:

```r
# cluster and order from scratch, ignoring existing linkage groups
mstmap(cross, bychr = FALSE, p.value = 1e-12)

# re-order within existing linkage groups, never splitting them
mstmap(cross, bychr = TRUE, p.value = 2)
```

A `p.value` greater than 1 switches clustering off altogether, which is what
makes the second idiom work.

## Choosing `p.value`

`p.value` sets the threshold at which markers are split into separate linkage
groups, and the right value depends strongly on population size — larger
populations need a smaller one. Expect to try a few. `pValue()` plots the
relationship, so the choice can be made deliberately rather than by trial and
error.

```{r pvalue, fig.height = 5.5}
pValue(dist = c(25, 30, 35, 40), pop.size = 100:500, map.function = "kosambi")
```

For a population of around 300, a `p.value` of `1e-12` gives a threshold of
roughly 30 cM before markers are linked into the same group.

## Checking the result

`heatMap()` shows pairwise recombination fractions and LOD scores of linkage
together, each with its own legend. Consistent heat within a linkage group
means the markers near each other really are linked.

```{r heat, fig.height = 8, warning = FALSE}
heatMap(mapDH, lmax = 50)
```

## Where to go next

| Article | What it covers |
|---|---|
| [How the MSTmap algorithm works](mstmap-algorithm.html) | The minimum spanning tree formulation, clustering and marker ordering |
| [Constructing a linkage map](constructing-a-map.html) | Both construction functions and every MSTmap parameter |
| [Pulling and pushing markers](pulling-and-pushing.html) | Setting problematic markers aside and reintroducing them |
| [Diagnosing genotypes and markers](diagnostics.html) | Profiling individuals, markers and intervals; genetic clones |
| [Heat maps](heat-maps.html) | Reading and tuning the recombination fraction and LOD display |
| [Manipulating linkage maps](manipulating-maps.html) | Breaking, merging, subsetting and combining maps |
| [Worked example I: construction](worked-example-construction.html) | A real barley backcross, from raw markers to a constructed map |
| [Worked example II: refinement](worked-example-refinement.html) | Pushing markers back, merging groups, fine mapping |
| [Notes on the MSTmap algorithm](algorithm-notes.html) | Distance calculations, `mvest.bc` and `detectBadData` |

If you use the package in published work, please cite it — `citation("ASMap")`,
or Taylor and Butler (2017), *Journal of Statistical Software* **79**(6), 1–29,
[doi:10.18637/jss.v079.i06](https://doi.org/10.18637/jss.v079.i06).
