---
title: "Heat maps"
description: >
  Displaying pairwise recombination fractions and LOD scores of linkage together, with a separate legend for each.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```

## Improved heat map

A visual diagnostic that is very useful for checking how well a linkage map is constructed is a heat map that combines the estimates of the pairwise recombination fraction (RF) between markers as well as LOD scores reflecting the the strength of linkage between each pair. Let the estimate of RF between marker $i$ and $j$ be $r_{ij}$ then the LOD score is a test of no linkage ($r_{ij} = 0.5$). Higher LOD scores indicate that the hypothesis of no linkage is rejected and the strength of the connection between the pair of markers is strong.

In R/qtl users can plot the heat map of pairwise linkage between markers using `plotRF()`. The `plotRF()` version of `mapDH` is given in the figure below. In this plot strong pairwise linkages between markers are represented as “hot” or red areas and weak pairwise linkages between markers are displayed as “cold” or blue areas.

```{r plotrffig, fig.width = 9, fig.height = 8, fig.cap = "Heat map of the constructed linkage map mapDH using `plotRF()`.", fig.alt = "Heat map of the constructed linkage map mapDH using `plotRF()`."}
plotRF(mapDH)
```

Unfortunately the plot contains some inadequacies that are mostly pointed out in the documentation (see `?plotRF`). “Recombination fractions are transformed by -4(log2(r)+1) to make them on the same sort of scale as LOD scores. Values of LOD or the transformed recombination fraction that are above 12 are set to 12.” This transformation and the arbitrary threshold induce multiple restrictions on the visualization of the heat map. Firstly, the LOD score is highly dependent on the number of individuals in the population and the upper limit of 12 may be a vastly inadequate representation of the pairwise linkage between some markers. In the figure below this latter inadequacy is displayed as red or hot areas of linkage between markers that should otherwise be cooler colours. Additionally, the actual estimated recombination fractions are not displayed in the upper triangle of the heat map and colour legends matching the numerical LOD and RF’s are not present. From a visual standpoint, it is well known the rainbow colour spectrum is perceptually inadequate for representing diverging numerical data.

R/ASMap contains an improved version of the heat map called `heatMap()` that rectifies the numerical and visual issues of the **qtl** heat map.

```r
heatMap(x, chr, mark, what = c("both", "lod", "rf"), lmax = 12,
             rmin = 0, markDiagonal = FALSE, color = rev(rainbow(256, start =
             0, end = 2/3)), ...)
```

The function independently plots the LOD score on the bottom triangle of the heatmap as well as the actual estimated recombination fractions on the upper triangle. The function also releases the arbitrary threshold on the LOD score and recombination fractions by providing a user defined `lmax` and `rmin` argument that is adjustable to suit the population size and any requirements of the plot required. For example, the figure below shows the `heatMap()` equivalent of `plotRF()` for the linkage map `mapDH` made with the call

```{r heat2, fig.cap = "Heat map of the constructed linkage map mapDH using `heatMap()`.", fig.alt = "Heat map of the constructed linkage map mapDH using heatMap()."}
heatMap(mapDH, lmax = 50)
```

The instantly visual difference between the figures is the chosen colour palette. R/ASMap uses a diverging colour palette `"Spectral"` from the default color palettes of **RColorBrewer**. The palette softens the plot and provides a gentle dinstinction between strongly linked and weakly linked markers. The plot also includes a legend for the LOD score and recombination fractions on either side of the figure below. As the estimated recombination fractions are being freely plotted without transformation, the complete scale is included in the heat map. This scale also includes estimated recombination fractions that are above the theoretical threshold of 0.5. By increasing this scale beyond 0.5, potential regions where markers out of phase with other markers can be recognised. The key to obtaining an “accurate” heat map is to match the heat of the LOD scores to the heat of the estimated recombination fractions and setting `lmax = 50` achieves this goal. Similar to `plotRF()`, the `heatMap()` function allows subsetting of the linkage map by `chr`. In addition, users can further subset the linkage groups using the argument `mark` by indexing a set of markers within linkage groups defined by `chr`.

