---
title: "Heat maps"
description: >
  Displaying pairwise recombination fractions and LOD scores of linkage
  together, with an independent legend for each.
bibliography: ../ref.bib
link-citations: true
nocite: |
  @tb17
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

The most informative single check on a constructed linkage map is a heat map
combining the estimated pairwise recombination fractions between markers with
the LOD scores expressing the strength of linkage between each pair. Let
$r_{ij}$ denote the estimated recombination fraction between markers $i$ and
$j$; the LOD score is then a test of no linkage, $r_{ij} = 0.5$. High LOD scores
indicate rejection of that hypothesis and therefore strong linkage between the
pair.

## The R/qtl display and its limitations

R/qtl provides `plotRF()` for this purpose [@br14]. Strong pairwise linkages
appear as hot, or red, areas and weak linkages as cold, or blue, areas.

```{r plotrffig, fig.width = 9, fig.height = 8, fig.cap = "Heat map of the constructed linkage map mapDH using `plotRF()`.", fig.alt = "Heat map of the constructed linkage map mapDH using plotRF()."}
plotRF(mapDH)
```

The display has several limitations, most of which are noted in its own
documentation: recombination fractions are transformed by $-4(\log_2(r)+1)$ to
bring them onto a scale comparable with LOD scores, and values of either
quantity exceeding 12 are set to 12.

This transformation and the fixed threshold impose three restrictions.

1. The LOD score depends strongly on the number of individuals in the
   population, so an upper limit of 12 may represent the pairwise linkage
   between some markers very poorly. In the figure above this appears as hot
   areas of linkage between markers that should be rendered in cooler colours.
2. The estimated recombination fractions are not themselves displayed in the
   upper triangle, and no legends relate the colours to the numerical LOD scores
   and recombination fractions.
3. The rainbow colour spectrum is perceptually unsuited to the representation of
   diverging numerical data.

## The ASMap display

[`heatMap()`](../reference/heatMap.html) addresses each of these. LOD scores are
plotted on the lower triangle and the estimated recombination fractions,
untransformed, on the upper. The fixed threshold is replaced by the user-defined
arguments `lmax` and `rmin`, which may be set to suit the population size and
the requirements of the plot.

```{r heat2, fig.height = 8, fig.cap = "Heat map of the constructed linkage map mapDH using `heatMap()`.", fig.alt = "Heat map of the constructed linkage map mapDH using heatMap()."}
heatMap(mapDH, lmax = 50)
```

The most immediate difference is the colour palette. ASMap uses the diverging
`"Spectral"` palette of **RColorBrewer**, which softens the display and draws a
gentler distinction between strongly and weakly linked markers. Separate legends
for the LOD scores and the recombination fractions appear on either side of the
plot.

Because the recombination fractions are plotted without transformation, the
complete scale is shown, including values above the theoretical threshold of
0.5. Extending the scale beyond 0.5 in this way allows regions in which markers
are out of phase with their neighbours to be recognised.

## Matching the two scales

> An accurate heat map is obtained when the heat of the LOD scores matches the
> heat of the estimated recombination fractions. Setting `lmax = 50` achieves
> this for `mapDH`; a different population size will require a different value.

This matching is the reason `lmax` exists as an argument rather than a constant.
Where the two triangles disagree visibly, the LOD threshold rather than the map
is usually at fault.

## Subsetting the display

As with `plotRF()`, the map may be subset by linkage group through `chr`. The
`mark` argument subsets further, indexing a set of markers within the linkage
groups nominated by `chr`, which is useful for examining a suspected junction
between two groups at close range.

## Further reading

- [Diagnosing genotypes and markers](diagnostics.html) — numerical counterparts
  to this graphical check
- [Worked example II](worked-example-refinement.html) — heat maps used to
  identify linkage groups requiring merging
- [`heatMap()`](../reference/heatMap.html) — full argument documentation

## References
