---
title: "Manipulating linkage maps"
description: >
  Breaking, merging, subsetting and combining linkage maps, and fast estimation of genetic distances.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```

## Miscellaneous additional R/qtl functions

### Breaking and merging linkage groups

During the linkage map construction process there may be a requirement to break or merge linkage groups. R/ASMap provides two functions to achieve this.

```r
breakCross(cross, split = NULL, suffix = "numeric", sep = ".")
mergeCross(cross, merge = NULL, gap = 5)
```

The `breakCross()` function allows users to break linkage groups in a variety ways. The `split` argument takes a list with elements named by the linkage group names that require splitting and containing the markers that immediately proceed where the splits are to be made. For example, a split of the linkage group 3B and 6A linkage map `mapDH` after the seventh and fifteenth marker respectively can be easily made using

```{r mb2}
mapDHb1 <- breakCross(mapDH, split = list("3B" = "3B.m.7","6A" = "6A.m.15"))
nmar(mapDHb1)
```

The `split` argument is flexible and can handle multiple linkage groups as well as multiple markers within linkage groups. The default use of the `suffix` argument produces a numerical suffix attachment to the original linkage groups being split with a separated by `sep`. Users can also provide their own complete names for the new split linkage groups by explicitly naming them in the `suffix` argument.

```{r mb3}
mapDHb2 <- breakCross(mapDH, split = list("3B" = "3B.m.7"), suffix = list("3B" = c("3B1","3B2")))
nmar(mapDHb2)
```

The `mergeCross()` function provides a method for merging linkage groups. Its argument `merge` requires a list with elements named by the proposed linkage group names required and containing the linkage groups to be merged. For the linkage map `mapDHb1` containing split linkage groups 3B and 6A created by a call to `breakCross()` the call to `mergeCross()` would be

```{r mb4}
mapDHm <- mergeCross(mapDHb1, merge = list("3B" = c("3B.1","3B.2"),"6A" = c("6A.1","6A.2")))
nmar(mapDHm)
```

It should be noted that this function places an artificial genetic distance gap between the merged linkage groups set by the `gap` argument. Accurate distance estimation would require a separate map estimation procedure after the merging has taken place.

### Rapid genetic distance estimation

The linkage map estimation function in R/qtl called `est.map()` can be invoked individually or can be applied through `read.cross()` when setting the argument `estimate.map = TRUE`. The function applies the multi-locus hidden Markov model technology of [@lg87] to perform its calculations. Unfortunately this technology is computationally cumbersome if there are many markers on a linkage group and becomes more so if there are many missing allele calls and genotyping errors present.

R/ASMap contains a small map estimation function that circumvents this computational burden.

```r
quickEst(object, chr, map.function = "kosambi", ...)
```

The function makes use of another function in R/qtl called `argmax.geno()`. This function is also a multi-locus hidden Markov algorithm that uses the observed markers present in a linkage group to impute pseudo-markers at any chosen cM genetic distance. In this case, we only require a reconstruction or imputation at the markers themselves. For the most accurate imputation to occur there needs to be an estimate of genetic distance already in place. To obtain an initial estimate of distance `est.rf()` is called within each linkage group defined by `chr` and recombination fractions are converted to genetic distances based on `map.function`. Unlike `est.map()` the `quickEst()` function lives up to its namesake by providing the quickest genetic distance calculations for large linkage maps.

```{r quick2, fig.width = 7, fig.height = 5, fig.cap = "Comparison of `mapDH` using `est.map` and `quickEst`."}
map1 <- est.map(mapDH, map.function = "kosambi")
map1 <- subset(map1, chr = names(nmar(map1))[6:15])
map2 <- quickEst(mapDH, map.function = "kosambi")
map2 <- subset(map2, chr = names(nmar(map2))[6:15])
plotMap(map1, map2)
```

The linkage map `mapDH` was re-estimated using `est.map()` and `quickEst()` and a comparison of the resulting maps are given in the figure below. The graphic indicates that there negligible changes in marker placement and overall linkage group distances between the two linkage maps.

### Subsetting in R/ASMap

The functions `pullCross()` and `pushCross()` described in section Pulling and pushing markers are used to create and manipulate extra list elements `"co.located"`, `"seg.distortion"` and `"missing"` associated with different marker types. Each element contains a data element consisting of a marker matrix equivalent in row dimension to the marker elements of the linkage map they were pulled from. Unfortunately, these list elements are not recognized by the native R/qtl functions. If the R/qtl function `subset.cross()` is used to subset the object to a reduced number of individuals then the data component of each of these elements will not be subsetted accordingly. In addition, the statistics in the table component of the elements `"seg.distortion"` and `"missing"` will be incorrect for the newly subsetted linkage map.

The `subsetCross()` function available in the R/ASMap package contains identical functionality to `subset.cross` but also ensures the data components of the extra list elements `"co.located"`, `"seg.distortion"` and `"missing"` are subsetted to match the linkage map. In addition, for elements `"seg.distortion"` and `"missing"` it also updates the table components to reflect the newly subsetted map. This update ensures that `pushCross()` uses the most accurate information when deciding which markers to push back into the linkage map.

Using the default `seg.thresh = 0.05` for the linkage map `mapDH`, distorted markers are pulled from the map

```{r sc}
mapDH.s <- pullCross(mapDH, type = "seg.distortion")
mapDH.s <- subsetCross(mapDH.s, ind = 3:218)
dim(mapDH.s$seg.distortion$data)[1]
```

In this example the use of `subsetCross()` ensures that the data component of the `"seg.distortion"` element is the same dimension as the map. The table element is also updated to ensure the statistics are correct for the reduced subset of lines.

### Combining maps

Over the period of time that I have been involved in linkage map construction there has been many occasions where I have required a function that could merge R/qtl cross objects together in an intelligent manner. For example, the merging of two linkage maps from the same population that were independently built with markers from two different platforms. This idea was the motivation behind the `combineMap()` function in the R/ASMap package. The aim of the function was to merge linkage maps based on map information, readying the combined linkage groups for reconstruction through an efficient linkage map construction process such as `mstmap.cross()`.

```r
combineMap(..., id = "Genotype", keep.all = TRUE)
```

The function takes an unlimited number of maps through the `...` argument. The linkage maps must all have the same cross class structure and contain the same genotype identifier `id`. At the current stage of writing this vignette the function required unique maker names across all linkage maps. This is expected to be relaxed at a later date so linkage maps that share markers, such as nested association mapping populations, can be merged effectively.

The merging of the maps happens intelligently using several components of the map. Firstly the linkage maps are merged based on commonality between the genotypes. If `keep.all = TRUE` the new combined linkage map is “padded out” with missing values where genotypes are not shared. If `keep.all= FALSE` the combined map is reduced to genotypes that are shared among all linkage maps. Secondly, if linkage group names are shared between maps then the markers from the shared linkage groups are merged. To exemplify its use a duplicate of `mapDH` is made and 10 linkage group names have been altered, with the marker names inside each of the linkage groups also altered to ensure they are unique.

```{r comb2}
mapDH1 <- mapDH
names(mapDH1$geno)[5:14] <- paste("L",1:10, sep = "")
mapDH1$geno <- lapply(mapDH1$geno, function(el){
  names(el$map) <- dimnames(el$data)[[2]] <- paste(names(el$map), "A", sep = "")
  el})
mapDHc <- combineMap(mapDH, mapDH1)
nmar(mapDHc)
```

The resulting combined map includes a combined marker set for the linkage groups that shared the same name and distinct linkage groups for unshared names. Again, note the function only merges the linkage maps and does not reconstruct the final combined linkage map.

The advantages of this function may not be obvious at a first glance. If an attempt is made to completely reconstruct the super set of linkage maps, rather than combine them first, the identification of linkage groups is lost. This function serves to preserve the important identity of linkage groups. More examples of its use in common map construction procedures will be explored in section chapter Post-construction linkage map development of the next chapter.

