---
title: "Manipulating linkage maps"
description: >
  Breaking, merging, combining and subsetting linkage maps, and rapid
  estimation of genetic distances.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

Construction is rarely a single operation. Linkage groups may need to be broken
or merged, maps built on different marker platforms may need to be combined, and
individuals may need to be removed. The functions described here perform these
operations on an R/qtl cross object while preserving the additional structures
ASMap places within it.

| Function | Operation |
|---|---|
| [`breakCross()`](../reference/breakCross.html) | Split linkage groups at nominated markers |
| [`mergeCross()`](../reference/mergeCross.html) | Combine nominated linkage groups into one |
| [`combineMap()`](../reference/combineMap.html) | Merge separate cross objects into a single map |
| [`subsetCross()`](../reference/subsetCross.html) | Subset genotypes or linkage groups consistently |
| [`quickEst()`](../reference/quickEst.html) | Estimate genetic distances rapidly |

## Breaking and merging linkage groups

The `split` argument of `breakCross()` takes a list whose elements are named by
the linkage groups to be split, and which contain the markers **immediately
preceding** each split point. Splitting 3B after its seventh marker and 6A after
its fifteenth is therefore expressed as follows.

```{r mb2}
mapDHb1 <- breakCross(mapDH, split = list("3B" = "3B.m.7", "6A" = "6A.m.15"))
nmar(mapDHb1)
```

Multiple linkage groups, and multiple split points within a group, may be
handled in a single call. By default a numeric suffix is appended to the name of
the group being split, separated by `sep`. Complete names for the new groups may
be supplied instead by naming them explicitly in `suffix`.

```{r mb3}
mapDHb2 <- breakCross(mapDH, split = list("3B" = "3B.m.7"),
                      suffix = list("3B" = c("3B1", "3B2")))
nmar(mapDHb2)
```

`mergeCross()` performs the converse operation. Its `merge` argument takes a
list whose elements are named by the proposed linkage group names and which
contain the groups to be merged.

```{r mb4}
mapDHm <- mergeCross(mapDHb1, merge = list("3B" = c("3B.1", "3B.2"),
                                           "6A" = c("6A.1", "6A.2")))
nmar(mapDHm)
```

> `mergeCross()` inserts an artificial genetic distance gap between the merged
> groups, governed by the `gap` argument. The distances across the join are
> therefore not estimates, and a separate estimation step is required after
> merging.

## Combining maps

A recurring requirement in linkage map construction is the merging of cross
objects, for instance where two maps of the same population have been built
independently from markers on different platforms. `combineMap()` addresses
this, merging maps on the basis of their map information so that the combined
linkage groups are ready for reconstruction by `mstmap.cross()`.

Any number of maps may be supplied. They must share the same cross class
structure and the same genotype identifier `id`, and marker names must be unique
across all of the maps supplied.

Merging proceeds in two stages. Maps are first merged on the genotypes they have
in common: with `keep.all = TRUE` the combined map is padded with missing values
where genotypes are not shared, whereas with `keep.all = FALSE` it is reduced to
those genotypes common to every map. Linkage groups bearing the same name across
maps are then merged, and groups whose names are not shared are retained
separately.

The example below duplicates `mapDH`, renames ten of its linkage groups and
alters the marker names within each group to ensure uniqueness.

```{r comb2}
mapDH1 <- mapDH
names(mapDH1$geno)[5:14] <- paste("L", 1:10, sep = "")
mapDH1$geno <- lapply(mapDH1$geno, function(el) {
  names(el$map) <- dimnames(el$data)[[2]] <- paste(names(el$map), "A", sep = "")
  el
})
mapDHc <- combineMap(mapDH, mapDH1)
nmar(mapDHc)
```

The combined map holds a merged marker set for those linkage groups that shared
a name, and distinct groups for those that did not.

> `combineMap()` merges maps but does not reconstruct the result, and it applies
> no consensus map algorithm. Chromosome identity and genetic distances are
> taken from the first appearance of a marker across the maps supplied. Run
> `mstmap()` afterwards.

The value of combining before reconstructing is not immediately obvious.
Reconstructing the union of several maps directly would discard the identity of
the linkage groups; combining first preserves it. Applications of this in
practice are given in
[Worked example II](worked-example-refinement.html).

## Subsetting

`pullCross()` and `pushCross()` create the additional list elements
`"co.located"`, `"seg.distortion"` and `"missing"`, each holding a marker matrix
whose row dimension matches the marker elements of the map it was drawn from.
These elements are not known to R/qtl.

> Using `subset.cross()` on an object carrying those elements will subset the
> map but **not** the elements within it. Their data components will retain the
> removed individuals, and the statistics in the tables of the
> `"seg.distortion"` and `"missing"` elements will no longer describe the map.
> Nothing reports this.

`subsetCross()` provides the functionality of `subset.cross()` while subsetting
those data components to match, and additionally recalculates the table
components of the `"seg.distortion"` and `"missing"` elements. That
recalculation matters, because `pushCross()` consults those statistics when
deciding which markers to restore.

```{r sc}
mapDH.s <- pullCross(mapDH, type = "seg.distortion")
mapDH.s <- subsetCross(mapDH.s, ind = 3:218)
dim(mapDH.s$seg.distortion$data)[1]
```

The data component of the `"seg.distortion"` element now has the same dimension
as the map.

## Rapid genetic distance estimation

The map estimation function of R/qtl, `est.map()`, applies the multi-locus
hidden Markov model of @lg87. That approach is computationally demanding for
linkage groups carrying many markers, and increasingly so in the presence of
missing allele calls and genotyping errors.

`quickEst()` circumvents this by a different route. It calls `argmax.geno()`
from R/qtl, itself a multi-locus hidden Markov algorithm, but requires
reconstruction only at the markers themselves rather than at imputed
pseudo-markers. Accurate imputation requires an initial estimate of genetic
distance, which is obtained by calling `est.rf()` within each linkage group
nominated by `chr` and converting the recombination fractions to distances
according to `map.function`.

```{r quick2, fig.width = 7, fig.height = 5, fig.cap = "Comparison of `mapDH` estimated using `est.map()` and using `quickEst()`.", fig.alt = "Comparison of mapDH estimated using est.map() and using quickEst()."}
map1 <- est.map(mapDH, map.function = "kosambi")
map1 <- subset(map1, chr = names(nmar(map1))[6:15])
map2 <- quickEst(mapDH, map.function = "kosambi")
map2 <- subset(map2, chr = names(nmar(map2))[6:15])
plotMap(map1, map2)
```

The comparison shows negligible differences in marker placement and in overall
linkage group distances between the two methods.

## Further reading

- [Pulling and pushing markers](pulling-and-pushing.html) — the elements that
  make `subsetCross()` necessary
- [Worked example II](worked-example-refinement.html) — merging linkage groups
  and combining maps in practice
- [Constructing a linkage map](constructing-a-map.html) — reconstruction after
  any of these operations

## References
