---
title: "How the MSTmap algorithm works"
description: >
  The minimum spanning tree formulation underlying ASMap: the clustering of
  markers into linkage groups, the determination of an optimal order within a
  group, and the imputation of missing scores and detection of genotyping
  errors that are integrated with it.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

This article describes the technical basis of the MSTmap algorithm [@mst08] as
it is implemented in ASMap. An understanding of its components is of practical
value, since they correspond directly to arguments of the two construction
functions.

## Background

Genetic linkage maps consist of a set of polymorphic markers spanning the
genome of a population derived from a cross between parental lines. They are
used both to examine the genetic landscape of the population itself and, more
commonly, to conduct gene–trait association studies such as quantitative trait
loci (QTL) analysis or genomic selection. In QTL analysis in particular, the
interpretation of significant genomic locations is improved when the markers
have been assigned to chromosomal groups and optimally ordered within them.

Within the R environment [@rsoft14] a small number of packages perform linkage
map construction. R/qtl [@br14] provides an extensive set of tools for
construction, exploration and manipulation across backcross, doubled haploid,
intercrossed F2 and four-way populations, together with advanced recombinant
inbred lines and populations arising from repeated backcrossing and selfing; it
is documented at length by @brosen09. R/onemap [@one14] offers comparable
facilities for backcross, doubled haploid and recombinant inbred populations,
and additionally for outcrossing populations.

Marker ordering in both packages proceeds in two stages. An initial order is
obtained by a non-exhaustive method such as SERIATION [@ser87] or RECORD
[@rec05]; because such an order is not guaranteed to be optimal, it is then
refined by exhaustively evaluating all permutations within a sliding marker
window, a procedure known in the linkage mapping community as *rippling*. The
computational cost of the second stage grows rapidly with the number of markers
and the width of the window, and the procedure may require repetition before a
stable order is reached.

ASMap was developed to avoid this two-stage structure [@tb14; @tb17]. Its
construction functions use the MSTmap algorithm [@mst08], implemented as C++
source code and available separately from
<http://alumni.cs.ucr.edu/~yonghui/mstmap.html>. The algorithm uses the minimum
spanning tree of a graph [@ct76] both to cluster markers into linkage groups
and to order markers within each group, in a single stage. It is applicable to
backcross, doubled haploid and recombinant inbred populations; for the latter
the level of self pollination may be specified, so that selfed F2, F3, $\ldots$,
F*n* populations are accommodated, where $n$ denotes the level of selfing.
Advanced recombinant inbred populations are treated as bi-parental for the
purpose of construction.

## Notation and objective functions

Following the notation of @mst08, consider a doubled haploid population of $n$
individuals genotyped across a set of $t$ markers, in which each $(i,j)$th entry
of the $n \times t$ matrix $\boldsymbol{M}$ is either an $A$ or a $B$,
representing the two parental homozygotes. Let $\boldsymbol{P}_{jk}$ denote the
probability of a recombination event between the markers $(\boldsymbol{m}_j,
\boldsymbol{m}_k)$, where $0 \leq \boldsymbol{P}_{jk} < 0.5$. MSTmap admits two
weight objective functions based on these recombination probabilities,

$$\begin{aligned}
  w_p(j,k) &= \boldsymbol{P}_{jk} \qquad (1)\\
  w_{ml}(j, k) &= -\bigl(\boldsymbol{P}_{jk}\log\,\boldsymbol{P}_{jk} + (1 -
  \boldsymbol{P}_{jk})\log(\,1 - \boldsymbol{P}_{jk})\bigr) \qquad (2)
\end{aligned}$$

In general $\boldsymbol{P}_{jk}$ is unknown and is replaced by the estimate
$d_{jk}/n$, where $d_{jk}$ is the Hamming distance between $\boldsymbol{m}_j$
and $\boldsymbol{m}_k$, that is, the number of non-matching alleles between the
two markers. This estimate is also the maximum likelihood estimate of
$\boldsymbol{P}_{jk}$ for both weight functions.

These correspond to the `objective.fun` argument of the construction functions,
`"COUNT"` selecting (1) and `"ML"` selecting (2).

## Clustering markers into linkage groups

If markers $\boldsymbol{m}_j$ and $\boldsymbol{m}_k$ belong to different linkage
groups then $\boldsymbol{P}_{jk} = 0.5$, and the Hamming distance between them
has the property $E(d_{jk}) = n/2$. A thresholding mechanism for determining
whether two markers belong to the same linkage group follows from Hoeffding's
inequality,

$$\begin{aligned}
 P(d_{jk} < \delta) \leq \mbox{exp}(-2(n/2 - \delta)^2/n) \qquad (3)
\end{aligned}$$

for $\delta < 0.5$. For a given $P(d_{jk} < \delta) = \epsilon$ and $n$, the
equation $-2(n/2 - \delta)^2/n = \mbox{log}\,\epsilon$ is solved to obtain an
appropriate Hamming distance threshold $\hat{\delta}$.

@mst08 indicate that the choice of $\epsilon$ is not critical to the formation
of linkage groups. The equation to be solved is, however, strongly dependent on
the number of individuals in the population, and in practice the choice matters
a great deal. The figure below shows, for a doubled haploid population, the
profiles of $-\log_{10} \epsilon$ against population size for four threshold
minimum cM distances $(25, 30, 35, 40)$.

```{r pvalfig, fig.width = 8, fig.height = 5.5, fig.cap = "Negative log10 epsilon versus the number of genotypes in the population, for four threshold cM distances.", fig.alt = "Negative log10 epsilon versus the number of genotypes in the population, for four threshold cM distances."}
pValue(dist = c(25, 30, 35, 40), pop.size = 100:500, map.function = "kosambi")
```

MSTmap adopts a default of $\epsilon = 0.00001$, which is generally suitable for
population sizes of $n \sim 150$ to $200$. For larger populations the default
becomes unsuitable. At $n = 350$, for instance, the figure indicates that
$\epsilon$ between $1.0e^{-12}$ and $1.0e^{-15}$ imposes a conservative minimum
threshold of 30 to 35 cM before markers in different clusters are linked,
whereas retaining the default lowers that threshold to approximately 45 cM, at
which point distinct clusters of markers will appear linked.

> The relationship between population size and `p.value` should be examined
> before construction is attempted. It is the argument that most often accounts
> for an implausible number of linkage groups.

Clustering itself proceeds by forming an edge-weighted undirected complete graph
for $\boldsymbol{M}$, in which the markers are vertices and the edge between any
two markers $\boldsymbol{m}_j$ and $\boldsymbol{m}_k$ carries the pairwise
Hamming distance $d_{jk}$. Edges of weight greater than $\hat{\delta}$ are
removed, and the remaining connected components partition the marker set into
$r$ linkage groups, $\boldsymbol{M} = [\boldsymbol{M}_1,\ldots,\boldsymbol{M}_r]$.

## Ordering markers within a linkage group

Consider now the $n \times t$ matrix of markers $\boldsymbol{M}$ belonging to a
single linkage group. Prior to ordering, the markers are binned into groups
within which the pairwise distance between any two markers is zero. Markers
within such a group exhibit no recombination between them and are said to
co-locate at the same genomic position for the $n$ genotypes used in
construction. A representative marker is selected from each bin to form the
reduced $n \times t^*$ marker set $\boldsymbol{M}^*$.

For the reduced matrix, consider the complete set of entries $(j, k) \in (1,
\ldots, t^*)$ under either weight function (1) or (2). These may be viewed as the
upper triangle of a symmetric weight matrix $\boldsymbol{W}$, and equivalently
as the connected edges of an undirected graph in which the markers are vertices.
An order for the set $\boldsymbol{M}^*$, also termed a travelling salesman path,
is then obtained by visiting each marker once and summing the weights of the
edges traversed. To find a path of minimum weight, MSTmap employs a minimum
spanning tree algorithm [@ct76] such as that of @prim57.

Where the minimum weight path is unique, the minimum spanning tree gives the
correct marker order. Where the data contain genotyping errors, or the number of
individuals is small, the tree may not form a complete path and may instead
contain markers or small sets of markers as nodes attached to it. In such cases
MSTmap takes the longest path in the tree as a backbone and applies local
optimisation techniques, among them K-opt, node relocation and block
optimisation [see @mst08], to improve upon the current path. The integration of
these techniques is what permits ordering to be completed in a single stage.

## Imputation of missing scores

Imputation of missing allele scores is performed by an EM type algorithm that
is tightly integrated with the ordering procedure rather than applied as a
preliminary step. The marker matrix $\boldsymbol{M}^*$ is converted to a matrix
$\boldsymbol{A}$ whose entries express the probabilistic certainty of an allele
being $A$. For the $j$th marker and $i$th individual,

$$\begin{aligned}
  \boldsymbol{A}(i,j) = \left\{\begin{array}{ll}
          1 & \quad \mbox{if $\boldsymbol{M}^*(i,j)$ is the A allele}\\
          0 & \quad \mbox{if $\boldsymbol{M}^*(i,j)$ is the B allele}\\
          \frac{(1 - \hat{\boldsymbol{P}}_{j - 1,j})(1 - \hat{\boldsymbol{P}}_{j,j+1})}
          {(1 - \hat{\boldsymbol{P}}_{j,j+1})(1 - \hat{\boldsymbol{P}}_{j,j+1})
           + \hat{\boldsymbol{P}}_{j-1,j} \hat{\boldsymbol{P}}_{j,j+1}} & \quad \mbox{if
           $\boldsymbol{M}^*(i,j)$ is missing} \end{array}\right.\qquad (4)
\end{aligned}$$

where $\hat{\boldsymbol{P}}_{j,j-1}$ and $\hat{\boldsymbol{P}}_{j,j+1}$ are the
estimated recombination fractions between the $(j-1)$th and $j$th markers and
between the $j$th and $(j+1)$th markers respectively. The third expression is
the posterior probability that the missing value at marker $j$ is the $A$ allele
for genotype $i$, given the current estimates.

The procedure begins by calculating pairwise normalised distances between all
markers in $\boldsymbol{M}^*$ and deriving an initial weight matrix
$\boldsymbol{W}$. An undirected graph is formed from the markers and the upper
triangular entries of $\boldsymbol{W}$, and its minimum spanning tree
establishes an initial order. For the current order at the $(j-1, j, j+1)$th
markers, the E-step updates the missing observation at marker $j$ through the
estimates $\hat{\boldsymbol{P}}_{j-1,j} = \hat{d}_{j-1,j}/n$ and
$\hat{\boldsymbol{P}}_{j,j+1} = \hat{d}_{j,j+1}/n$ in (4). The M-step then
re-estimates the pairwise distances between all markers, where for the $j$th and
$k$th marker

$$\begin{aligned}
\hat{d}_{jk} = \sum_{i = 1}^{t^*}\boldsymbol{A}(i,j)(1 - \boldsymbol{A}(i, k)) + \boldsymbol{A}(i,k)(1 - \boldsymbol{A}(i, j))
\qquad (5)
\end{aligned}$$

and $\boldsymbol{W}$ is recalculated. A new undirected graph is formed, a new
order derived from its minimum spanning tree, and the procedure repeated to
convergence. Several iterations are generally required, though the computation
remains expedient; computation time increases with the number of missing values.

This procedure underlies the `mvest.bc` argument, which is discussed further in
[Notes on the MSTmap algorithm](algorithm-notes.html).

## Detection of genotyping errors

The algorithm can additionally detect and remove probable genotyping errors,
again as part of the ordering procedure. A weighted average of nearby markers
determines the expected state of an allele; for individual $i$ and marker $j$,

$$\begin{aligned}
\mbox{E}[\boldsymbol{A}(i, j)] = \sum_{j \neq
    k}d_{j,k}^{-2}\boldsymbol{A}(i,k)\bigg/\sum_{j \neq k}d_{j,k}^{-2}
\qquad (6)
\end{aligned}$$

where the weights are the inverse squared distances from marker $j$ to its
neighbours. Only a small set of nearby markers is used at each iteration, and an
observed allele is regarded as suspicious when $|\mbox{E}[\boldsymbol{A}(i, j)]
- \boldsymbol{A}(i, j)| > 0.75$. A suspicious observation is set to missing and
imputed by the EM algorithm described above.

> Removing a suspicious allele reduces the number of recombinations between the
> marker concerned and its neighbours. This has a direct effect on the estimated
> genetic distances and hence on the overall length of the linkage group, and
> the option should therefore be used deliberately rather than routinely.

This is the basis of the `detectBadData` argument, examined in
[Notes on the MSTmap algorithm](algorithm-notes.html).

## Further reading

- [Constructing a linkage map](constructing-a-map.html) — the arguments through
  which the components described above are controlled
- [Notes on the MSTmap algorithm](algorithm-notes.html) — the practical
  behaviour of `mvest.bc` and `detectBadData`, and how genetic distances are
  calculated
- [Worked example I](worked-example-construction.html) — the algorithm applied
  to a barley backcross population

## References
