---
title: "Pulling and pushing markers"
description: >
  Setting problematic markers aside before construction and reintroducing them
  afterwards, in place of their permanent removal.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

Linkage map construction is commonly preceded by some pruning of the marker
set, typically the removal of markers exhibiting a high proportion of missing
values or significant departure from their expected Mendelian segregation
ratios. Such removal is usually permanent, and the potential value of the
discarded markers is thereby overlooked. A marker that appears unsuitable when
judged against unordered data may prove entirely well behaved once a map has
been established.

A preferable approach is to identify the problematic markers and place them
aside, with the intention of reassessing their usefulness at a later stage of
the construction process. The functions
[`pullCross()`](../reference/pullCross.html) and
[`pushCross()`](../reference/pushCross.html) provide this facility. Rather than
deleting a marker, `pullCross()` transfers it to a holding element of the cross
object, retaining the information required to restore it;
`pushCross()` subsequently reintroduces it into an established map.

The constructed map `mapDH` is used below to illustrate both operations.

## Marker types

Three types of marker may be set aside.

| `type` | Markers affected | Rationale |
|---|---|---|
| `"co.located"` | Markers separated from another marker by zero distance | They contribute no ordering information, yet add to the computational cost of construction |
| `"seg.distortion"` | Markers whose segregation distortion p-value falls below `seg.thresh` | Distorted markers may not map to a unique location |
| `"missing"` | Markers whose proportion of missing values exceeds `miss.thresh` | Sparsely scored markers are positioned less reliably |

The relevant thresholds are supplied by
[`pp.init()`](../reference/pp.init.html), which provides defaults of
`seg.thresh = 0.05` and `miss.thresh = 0.1`. These may be overridden through the
`pars` argument.

## Setting markers aside

In the following, markers are pulled that are co-located with another marker,
that exhibit segregation distortion at a p-value below 0.02, or that are
missing more than 3% of their scores.

```{r pp2}
mapDHs <- pullCross(mapDH, type = "co.located")
mapDHs <- pullCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.02))
mapDHs <- pullCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.03))

names(mapDHs)
```

The object has acquired one element for each marker type. Each contains a
`table` summarising the markers concerned and a `data` matrix of their scores,
in the same genotype by marker layout as the linkage groups themselves.

```{r pp3}
mapDHs$seg.distortion$table
```

For the types `"seg.distortion"` and `"missing"`, the table combines positional
information with the output of `geno.table()` from the R/qtl package [@br14],
so that the basis for each marker's removal may be inspected.

```{r pp4}
head(mapDHs$co.located$table)
```

The table for `"co.located"` markers is of a different form. Markers are
assigned to bins, and within each bin the **first marker is retained in the map
as a reference**, the remainder being pulled. It is this reference that permits
the co-locating markers to be restored to their correct position.

## Reconstruction

Reconstruction with `bychr = FALSE` combines all remaining markers and rebuilds
the map in its entirety, in the course of which the linkage groups are renamed.
This provides a reasonably stringent test of whether the markers set aside can
subsequently be restored to their correct positions.

```{r pp5}
mapDHs <- mstmap(mapDHs, bychr = FALSE, dist.fun = "kosambi", trace = FALSE,
                 anchor = TRUE)
nmar(mapDHs)
```

## Reversal of the thresholds

The thresholds operate in opposite senses in the two functions, and this is a
frequent source of confusion.

> `pullCross()` removes a marker when its segregation distortion p-value lies
> **below** `seg.thresh`, whereas `pushCross()` restores it when that p-value
> lies **above** `seg.thresh`. The same applies to `miss.thresh`.

Pulling and pushing at an identical threshold therefore returns no markers at
all, and the function reports that there are none of that type to push back.
The two thresholds are intended to differ, the value used when pushing being
the less stringent of the two, so that a marker is restored only if it is
adequately behaved in the context of the established map.

| Parameter | Pulled when | Restored when |
|---|---|---|
| `seg.thresh` | p-value **<** threshold | p-value **>** threshold |
| `miss.thresh` | missing proportion **>** threshold | missing proportion **<** threshold |

Markers were pulled above at `seg.thresh = 0.02` and `miss.thresh = 0.03`, and
are accordingly restored at 0.001 and 0.05 respectively.

```{r pp6}
mapDHs <- pushCross(mapDHs, type = "co.located")
mapDHs <- pushCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.001))
mapDHs <- pushCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.05))

names(mapDHs)
```

Under these settings all markers of each type have been restored, and the
holding elements have consequently been removed from the object.

## Placement of the restored markers

```{r pp7}
pull.map(mapDHs)[[4]]
pull.map(mapDHs)[[21]]
```

Co-located markers are placed adjacent to their reference marker, as for
example `1D.m.4` beside `1D.m.3` on 1D. Markers held in the
`"seg.distortion"` and `"missing"` elements cannot be positioned so precisely
and are appended to the end of the linkage group to which they are assigned;
the distorted marker `6D.m.12` appears at the end of 6D.

> `pushCross()` assigns markers to linkage groups but does not order them. The
> map is not complete until MSTmap has been run again.

A final pass within each linkage group places all markers in their optimal
position. Setting `p.value = 2` suppresses clustering, so that the composition
of the groups cannot change.

```{r pp8}
mapDHs <- mstmap(mapDHs, bychr = TRUE, dist.fun = "kosambi", trace = FALSE,
                 anchor = TRUE, p.value = 2)
```

## Markers not previously present in the map

`pushCross()` admits a fourth type, `"unlinked"`, which distributes markers
residing in an unlinked group of the `geno` element into the established
linkage groups. This provides the mechanism by which new markers are introduced
into a completed map, and is described in
[Worked example II](worked-example-refinement.html) in the context of fine
mapping and of unknown linkage groups.

## Further reading

- [Diagnosing genotypes and markers](diagnostics.html) — determining which
  markers warrant setting aside
- [Worked example I](worked-example-construction.html) — the pull operation
  within a complete construction
- [`pullCross()`](../reference/pullCross.html),
  [`pushCross()`](../reference/pushCross.html),
  [`pp.init()`](../reference/pp.init.html) — full argument documentation

## References
