---
title: "Pulling and pushing markers"
description: >
  Setting problematic markers aside before construction and reintroducing them afterwards, instead of discarding them permanently.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH)
```

Most map construction begins by pruning markers — dropping those with too many
missing scores, or those badly distorted from their expected Mendelian ratios.
That pruning is usually permanent, and it throws away information you cannot get
back. A marker that looks unusable against raw data may be perfectly well
behaved once the map exists.

[`pullCross()`](../reference/pullCross.html) offers the alternative. Instead of
deleting a marker it moves it into a holding element of the cross object,
keeping everything needed to put it back.
[`pushCross()`](../reference/pushCross.html) then reintroduces it once you have
a map to judge it against.

This article pulls markers out of the constructed map `mapDH`, rebuilds it, and
pushes them back.

## Three kinds of problem marker

| `type` | Markers affected | Why set them aside |
|---|---|---|
| `"co.located"` | Markers at zero distance from another marker | They add no ordering information but cost construction time |
| `"seg.distortion"` | Markers whose segregation distortion p-value is below `seg.thresh` | Distorted markers may not map to a unique location |
| `"missing"` | Markers whose missing proportion exceeds `miss.thresh` | Sparse markers are placed less reliably |

Thresholds come from [`pp.init()`](../reference/pp.init.html), which supplies
defaults of `seg.thresh = 0.05` and `miss.thresh = 0.1`. Override them through
the `pars` argument.

## Pulling markers aside

Pull the co-located markers, those distorted at a p-value below 0.02, and those
missing more than 3% of their scores:

```{r pp2}
mapDHs <- pullCross(mapDH, type = "co.located")
mapDHs <- pullCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.02))
mapDHs <- pullCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.03))

names(mapDHs)
```

The object has gained one element per marker type. Each holds a `table`
summarising the markers and a `data` matrix of their scores, in the same
genotype-by-marker layout as the linkage groups themselves.

```{r pp3}
mapDHs$seg.distortion$table
```

For `"seg.distortion"` and `"missing"`, the table combines positional
information with the output of `geno.table()` from R/qtl, so you can see exactly
why each marker was pulled.

```{r pp4}
head(mapDHs$co.located$table)
```

The `"co.located"` table works differently. Markers are grouped into bins, and
within each bin the **first marker stays in the map as a reference**; the rest
are pulled. That reference is what lets `pushCross()` find its way back.

## Rebuild the map

Rebuilding with `bychr = FALSE` bulks every remaining marker and reconstructs
from scratch, which renames the linkage groups — a good test of whether the
markers really can find their way home.

```{r pp5}
mapDHs <- mstmap(mapDHs, bychr = FALSE, dist.fun = "kosambi", trace = FALSE,
                 anchor = TRUE)
nmar(mapDHs)
```

## The thresholds reverse when you push

This is the part that catches people out.

> `pullCross()` removes a marker when its distortion p-value is **below**
> `seg.thresh`. `pushCross()` returns it when that p-value is **above**
> `seg.thresh`. The same is true of `miss.thresh`, in the same direction.

So pulling and pushing at the same threshold returns nothing, and ASMap will
tell you there are no markers to push back. The two thresholds are meant to
differ, with the push set more permissively than the pull, so that a marker has
to earn its place in the map.

| Parameter | Pulled when | Pushed back when |
|---|---|---|
| `seg.thresh` | p-value **<** threshold | p-value **>** threshold |
| `miss.thresh` | missing proportion **>** threshold | missing proportion **<** threshold |

Here the markers were pulled at `seg.thresh = 0.02` and `miss.thresh = 0.03`, so
they are pushed back at 0.001 and 0.05 respectively.

```{r pp6}
mapDHs <- pushCross(mapDHs, type = "co.located")
mapDHs <- pushCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.001))
mapDHs <- pushCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.05))

names(mapDHs)
```

Every marker went back, so the holding elements have been removed from the
object entirely.

## Where the markers land

```{r pp7}
pull.map(mapDHs)[[4]]
pull.map(mapDHs)[[21]]
```

Co-located markers are placed next to their reference marker — `1D.m.4` sits
beside `1D.m.3`. Markers from the `"seg.distortion"` and `"missing"` elements
cannot be placed so precisely, so they are appended to the end of whichever
linkage group they belong to. `6D.m.12` is there at the end of 6D.

> `pushCross()` assigns markers to linkage groups but does **not** order them.
> The map is not finished until you run MSTmap again.

A final pass within each linkage group puts everything in its optimal position.
`p.value = 2` disables clustering, so the groups themselves cannot change:

```{r pp8}
mapDHs <- mstmap(mapDHs, bychr = TRUE, dist.fun = "kosambi", trace = FALSE,
                 anchor = TRUE, p.value = 2)
```

## Pushing markers that were never in the map

`pushCross()` accepts a fourth type, `"unlinked"`, which takes markers sitting in
an unlinked group of the `geno` element and distributes them into the
established linkage groups. This is the mechanism behind adding new markers to a
finished map — see
[Worked example II](worked-example-refinement.html) for fine mapping and
unknown linkage groups.

## Where next

- [Diagnosing genotypes and markers](diagnostics.html) — deciding which markers
  are worth pulling in the first place
- [Worked example I](worked-example-construction.html) — pulling markers as part
  of a full construction
- [`pullCross()`](../reference/pullCross.html),
  [`pushCross()`](../reference/pushCross.html),
  [`pp.init()`](../reference/pp.init.html) — full argument documentation
