---
title: "Pulling and pushing markers"
description: >
  Setting problematic markers aside before construction and reintroducing them afterwards, instead of discarding them permanently.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapDH); data(mapDHf); data(mapBC); data(mapBCu); data(mapF2)
```

## Pulling and pushing markers

Often in linkage map construction, some pruning of the markers occurs before initial construction. This may be the removal of markers with a proportion of missing values higher than some desired threshold as well as markers that are significantly distorted from their expected Mendelian segregation patterns. Often the removal is permanent and the possible importance of some of these markers is overlooked. A preferable system would be to identify and place the problematic markers aside with the intention of checking their usefulness at a later stage of the construction process.

The R/ASMap package contains two functions that allow you to do this. The first is the function `pullCross()` which provide users with a mechanism to “pull” markers of certain types from a linkage map and place them aside. The complementary function `pushCross()` then allows users to “push” markers back into the linkage map at any stage of the construction process. These can be seen as helper functions for more efficient construction of linkage maps (see `?pullCross` and `?pushCross` for complete details).

```r
pullCross(object, chr, type = c("co.located","seg.distortion","missing"),
               pars = NULL, replace = FALSE, ...)
pushCross(object, chr, type = c("co.located","seg.distortion","missing","unlinked"),
               unlinked.chr = NULL, pars = NULL, replace = FALSE, ...)
pp.init(seg.thresh = 0.05, seg.ratio = NULL, miss.thresh = 0.1, max.rf =
             0.25, min.lod = 3)
```

In the current version of the package, the helper functions share three types of markers that can be “pulled/pushed” from linkage maps. These include markers that are co-located with other markers, markers that have some defined segregation distortion and markers with a defined proportion of missing values. If the argument `type` is `"seg.distortion"` or `"missing"` then the initialization function `pp.init` is used to determine the appropriate threshold parameter setting (`seg.thresh`, `seg.ratio`, `miss.thresh`) that will be used to pull/push markers from the linkage map. Users can set their own parameters by setting the `pars` argument (see examples below). For each of the different types, `pullCross()` will pull markers from the map and place them in separate elements of the cross object. Within the elements, vital information is kept that can be accessed by `pushCross()` to push the markers back at a later stage of linkage map construction.

The function `pushCross()` also contains another marker type called `"unlinked"` which, in conjunction with the argument `unlinked.chr`, allows users to push markers from an unlinked linkage group in the `geno` element of the `object` into established linkage groups. This mechanism becomes vital, for example, when pushing new markers into an established linkage map. An example of this is presented in section Unknown linkage groups.

Again, the constructed linkage map `mapDH` will be used to showcase the power of `pullCross()` and `pushCross()`. Markers are pulled from the map that are co-located with other markers, have significant segregation distortion with a p-value less than 0.02 and have a missing value proportion greater than 0.03.

```{r pp2}
mapDHs <- pullCross(mapDH, type = "co.located")
mapDHs <- pullCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.02))
mapDHs <- pullCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.03))
names(mapDHs)
names(mapDHs$co.located)
```

The cross object now contains three new elements named by the marker types that are pulled from the map. In each of the elements there are two elements, a table of information for the markers that are pulled and the actual marker data in genotype by marker format (i.e. exactly the same as the data contained in the linkage groups themselves).

```{r pp3}
mapDHs$seg.distortion$table
```

The table element of each of the marker types `"missing"` and `"seg.distortion"` consists of a summary of positional information as well as information from `geno.table()` in the R/qtl package.

```{r pp4}
head(mapDHs$co.located$table)
```

The table element for the marker type `"co.located"` contains information on the markers that are co-located and the group or bin they belong to. In each group the first marker is the reference marker that remains in the linkage map and the remaining markers are pulled from the map and placed aside.

Suppose now the map is undergone a construction or re-construction process so that the linkage groups are artificially renamed. To do this we will re-run MSTmap with `bychr = FALSE`.

```{r pp5}
mapDHs <- mstmap(mapDHs, bychr = FALSE, dist.fun = "kosambi", trace = TRUE, anchor = TRUE)
nmar(mapDHs)
```

The markers can now be pushed back into the linkage map using `pushCross()`. The complete set of co.located markers are pushed back as well as markers that have significant segregation distortion with p-values greater than 0.001 and markers that have a missing value proportion less than 0.05.

```{r pp6}
mapDHs <- pushCross(mapDHs, type = "co.located")
mapDHs <- pushCross(mapDHs, type = "seg.distortion", pars = list(seg.thresh = 0.001))
mapDHs <- pushCross(mapDHs, type = "missing", pars = list(miss.thresh = 0.05))
names(mapDHs)
```

With the above parameter settings all markers from each marker type are pushed back into the map and the marker type elements are removed from the `object`.

```{r pp7}
pull.map(mapDHs)[[4]]
pull.map(mapDHs)[[21]]
```

For co-located markers the reference marker for each group is used as a guide to place the set of co-locating markers back into the linkage map adjacent to the reference marker. For example, the co-locating marker `1D.m.4` on 1D appears adjacent to its reference marker `1D.m.3`. Markers from the `"seg.distortion"` and `"missing"` elements are pushed back to the end of the linkage groups ready for the map to be re-constructed by chromosome. For example the distorted marker `6D.m.12` is pushed to the end of 6D. A final run of MSTmap within each linkage group will produce the desired map with all markers in their optimal position.

```{r pp8}
mapDHs <- mstmap(mapDHs, bychr = TRUE, dist.fun = "kosambi", trace = TRUE, anchor = TRUE, p.value = 2)
```

