---
title: "Worked example I: construction"
description: >
  A barley backcross marker set taken from 3,023 unassigned markers through
  pre-construction diagnosis to a constructed map of nine linkage groups.
bibliography: ../ref.bib
link-citations: true
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapBCu)
```

This article works through the construction of a linkage map for a barley
backcross population, comprising 3,023 markers genotyped on 326 individuals and
supplied as an unconstructed R/qtl object of class `"bc"`. The refinement of the
resulting map is continued in
[Worked example II](worked-example-refinement.html).

```{r ex1}
data(mapBCu, package = "ASMap")
```

## Pre-construction

Construction seldom consists of applying an algorithm to marker data as
supplied. It is prudent to work through a checklist beforehand, to establish
that the genotypes and markers being used are of the best available quality. The
following is a non-exhaustive but ordered list for an unconstructed marker set.

1. Check missing allele scores, both across markers for each genotype and
   across genotypes for each marker. A high proportion of missing information
   may indicate problems with the physical genotyping.
2. Check for genetic clones, or individuals sharing a high proportion of
   matching allelic information.
3. Check markers for excessive segregation distortion. Highly distorted markers
   may not map to a unique location.
4. Check markers for switched alleles. Such markers will not cluster or link
   well during construction, and it is preferable to repair their alignment
   beforehand.
5. Check for co-locating markers. For large maps it is computationally more
   efficient to omit markers co-located with others, at least temporarily.

### Missing allele scores

R/qtl provides a simple graphical check on the structure of missing scores
across both genotypes and markers.

```{r ex3, fig.cap = "Missing allele scores for the unconstructed map mapBCu.", fig.alt = "Missing allele scores for the unconstructed map mapBCu."}
plotMissing(mapBCu)
```

The darkest lines indicate genotypes carrying large amounts of missing data,
which may reflect poor physical genotyping; these should be removed before
proceeding. The markers are otherwise well scored across the range of
genotypes. `statGen()` identifies the genotypes concerned, which are then
removed using the usual R/qtl facilities.

```{r ex4}
sg <- statGen(mapBCu, bychr = FALSE, stat.type = "miss")
mapBC1 <- subset(mapBCu, ind = sg$miss < 1600)
```

### Genetic clones

Highly related individuals inflate the segregation distortion of markers, so a
course of action, whether removal of individuals or the formation of consensus
genotypes, should be settled before further diagnosis. `genClones()` identifies
and reports them; see
[Diagnosing genotypes and markers](diagnostics.html) for the function itself.

```{r ex5}
gc <- genClones(mapBC1, tol = 0.95)
gc$cgd
```

Thirteen groups share more than 0.95 of their alleles. The accompanying
statistics require interpretation rather than acceptance. The first group
contains three pairs matched on 1,620 markers or fewer, and those same pairs
have approximately 1,400 markers at which an allele was present for one
genotype and missing for the other. That is insufficient evidence of cloning,
and the group is therefore removed from the table before consensus genotypes are
formed for the remainder.

```{r ex6}
cgd <- gc$cgd[-c(1, 4, 5), ]
mapBC2 <- fixClones(mapBC1, cgd, consensus = TRUE)
levels(mapBC2$pheno[[1]])[grep("_", levels(mapBC2$pheno[[1]]))]
```

### Segregation distortion

Segregation distortion is the phenomenon whereby observed allelic frequencies at
a locus deviate from those expected under Mendelian genetics. It may arise from
laboratory processes, or in local genomic regions from underlying biological
mechanisms [@lytt91].

The extent of distortion, the allelic proportions and the proportion of missing
values may be displayed together with `profileMark()`. Setting
`crit.val = "bonf"` annotates markers whose p-value for the test of segregation
distortion falls below a family-wise Bonferroni adjusted level of 0.05 divided
by the number of markers.

```{r ex8, fig.width = 11.9, fig.height = 7, warning = FALSE, fig.cap = "For individual markers, the negative log10 p-value for the test of segregation distortion, the proportion of each contributing allele and the proportion of missing values.", fig.alt = "For individual markers, the negative log10 p-value for the test of segregation distortion, the proportion of each contributing allele and the proportion of missing values."}
profileMark(mapBC2, stat.type = c("seg.dist", "prop", "miss"), crit.val = "bonf",
            layout = c(1, 4), type = "l", cex = 0.5)
```

Numerous markers are significantly distorted, three of them highly so, and the
proportion of missing values does not exceed 20% for any marker. The three
extreme markers are dropped on the basis of their allele proportions.

```{r ex9}
mm <- statMark(mapBC2, stat.type = "marker")$marker$AB
mapBC3 <- drop.markers(mapBC2, c(markernames(mapBC2)[mm > 0.98],
                                 markernames(mapBC2)[mm < 0.2]))
```

### Setting markers aside

Without a constructed map the origin of the segregation distortion cannot be
determined. Using distorted markers indiscriminately may nonetheless create
problems during construction. The more prudent course is to set the distorted
markers aside, construct the map from the less problematic remainder, and then
reintroduce them to establish whether their effect is beneficial or deleterious.
`pullCross()` and `pushCross()` exist for this purpose; see
[Pulling and pushing markers](pulling-and-pushing.html). Here they are also used
to set aside markers with 10 to 20% missing values, and co-located markers.

```{r ex10}
mapBC3 <- pullCross(mapBC3, type = "missing", pars = list(miss.thresh = 0.1))
mapBC3 <- pullCross(mapBC3, type = "seg.distortion", pars = list(seg.thresh = "bonf"))
mapBC3 <- pullCross(mapBC3, type = "co.located")
names(mapBC3)
sum(ncol(mapBC3$missing$data), ncol(mapBC3$seg.dist$data),
    ncol(mapBC3$co.located$data))
```

A total of 847 markers have been set aside in their respective elements, leaving
2,173 markers from which the map is constructed.

## Construction

The curated marker data in `mapBC3` may now be passed to `mstmap()`.

```{r ex11}
mapBC4 <- mstmap(mapBC3, bychr = FALSE, trace = FALSE, dist.fun = "kosambi",
                 p.value = 1e-12)
chrlen(mapBC4)
```

Setting `bychr = FALSE` bulks the complete marker set and constructs from
first principles, clustering the markers into linkage groups and then ordering
the markers within each. For a population of 309 individuals, `p.value = 1e-12`
gives a threshold of approximately 30 cM before markers in distinct clusters are
linked; the relationship is set out in
[How the MSTmap algorithm works](mstmap-algorithm.html). The resulting map
contains nine linkage groups, each optimally ordered.

### Assessment by heat map

```{r ex12, fig.height = 8, fig.cap = "Heat map of the constructed linkage map mapBC4.", fig.alt = "Heat map of the constructed linkage map mapBC4."}
heatMap(mapBC4, lmax = 70)
```

As discussed in [Heat maps](heat-maps.html), an accurate display is obtained
when the heat of the recombination fractions on the upper triangle matches that
of the LOD scores, which `lmax = 70` achieves here. The heat is consistent
across the markers within each linkage group, indicating strong linkage between
neighbouring markers, and the groups appear distinctly clustered.

### Assessment by recombination rate

A successful heat map does not expose subtler problems. One of the key
characteristics of a well constructed map is an appropriate recombination rate
among the genotypes. For this population each line is considered independent
with an expected rate of approximately 14 across the genome, yet the map
contains linkage groups exceeding the theoretical cutoff of 200 cM, which
suggests genotypes with inflated rates.

```{r ex15, fig.width = 11.9, fig.height = 7, warning = FALSE, fig.cap = "For individual genotypes, the number of recombinations, double recombinations and missing values for mapBC4.", fig.alt = "For individual genotypes, the number of recombinations, double recombinations and missing values for mapBC4."}
pg <- profileGen(mapBC4, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
                 id = "Genotype", xo.lambda = 14, layout = c(1, 3), lty = 2,
                 cex = 0.7)
```

Seven lines have recombination rates above 20, and the same lines carry
excessive missing values.

> The offending genotypes are removed with `subsetCross()` rather than
> `subset.cross()`, so that the `"co.located"`, `"seg.distortion"` and
> `"missing"` elements are subset and their statistics recalculated alongside
> the map. Those statistics govern which markers are later restored.

```{r ex16}
mapBC5 <- subsetCross(mapBC4, ind = !pg$xo.lambda)
mapBC6 <- mstmap(mapBC5, bychr = TRUE, dist.fun = "kosambi", trace = FALSE,
                 p.value = 1e-12)
chrlen(mapBC6)
```

The linkage group lengths have fallen considerably, and the recombination rates
of the remaining genotypes may be confirmed to lie within acceptable limits.

### Marker and interval profiles

```{r ex18, fig.width = 11.9, fig.height = 8.4, warning = FALSE, fig.cap = "Marker profiles of the negative log10 p-value for the test of segregation distortion, allele proportions and the number of double crossovers, together with the interval profile of the number of recombinations between adjacent markers in mapBC6.", fig.alt = "Marker profiles of segregation distortion, allele proportions and double crossovers, with the interval profile of recombinations between adjacent markers in mapBC6."}
profileMark(mapBC6, stat.type = c("seg.dist", "prop", "dxo", "recomb"),
            layout = c(1, 5), type = "l")
```

The profiles confirm the success of the construction: no more than one double
crossover occurs at any marker, and very few in total. They also show the extent
of biological distortion that can arise within a linkage group.

Closer inspection of the segregation distortion and allele proportion panels
indicates that linkage group L.3 and the short group L.5 have profiles that
would join were the groups merged, and similarly for L.8 and L.9. That
possibility is pursued in
[Worked example II](worked-example-refinement.html).

## Further reading

- [Worked example II](worked-example-refinement.html) — reintroducing the
  markers set aside, merging linkage groups and post-construction development
- [Diagnosing genotypes and markers](diagnostics.html) — the diagnostic
  functions used above
- [Pulling and pushing markers](pulling-and-pushing.html) — the mechanism for
  setting markers aside

## References
