---
title: "Worked example II: refinement"
description: >
  Continuing the barley backcross example: reintroducing the markers held
  aside, merging linkage groups to the expected seven, and post-construction
  development including fine mapping.
bibliography: ../ref.bib
link-citations: true
nocite: |
  @tb17
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = NA,
  collapse = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  out.width = "100%",
  dpi = 110
)
library(ASMap)
data(mapBCu)
```

> This article continues the worked example. The code below rebuilds the map as
> it stood at the end of
> [Worked example I](worked-example-construction.html).

```{r preamble, include = FALSE}
sg <- statGen(mapBCu, bychr = FALSE, stat.type = "miss")
mapBC1 <- subset(mapBCu, ind = sg$miss < 1600)
gc <- genClones(mapBC1, tol = 0.95)
cgd <- gc$cgd[-c(1, 4, 5), ]
mapBC2 <- fixClones(mapBC1, cgd, consensus = TRUE)
mm <- statMark(mapBC2, stat.type = "marker")$marker$AB
mapBC3 <- drop.markers(mapBC2, c(markernames(mapBC2)[mm > 0.98],
                                 markernames(mapBC2)[mm < 0.2]))
mapBC3 <- pullCross(mapBC3, type = "missing", pars = list(miss.thresh = 0.1))
mapBC3 <- pullCross(mapBC3, type = "seg.distortion", pars = list(seg.thresh = "bonf"))
mapBC3 <- pullCross(mapBC3, type = "co.located")
mapBC4 <- mstmap(mapBC3, bychr = FALSE, trace = FALSE, dist.fun = "kosambi",
                 p.value = 1e-12)
pg <- profileGen(mapBC4, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
                 id = "Genotype", xo.lambda = 14, layout = c(1, 3), lty = 2,
                 cex = 0.7)
mapBC5 <- subsetCross(mapBC4, ind = !pg$xo.lambda)
mapBC6 <- mstmap(mapBC5, bychr = TRUE, dist.fun = "kosambi", trace = FALSE,
                 p.value = 1e-12)
```

## Reintroducing the markers set aside

The markers placed aside during pre-construction are now returned to the
constructed map, which is then re-diagnosed. The 515 markers carrying between
10% and 20% missing values, held in the `"missing"` element, are restored first.

```{r ex19}
mapBC6 <- pushCross(mapBC6, type = "missing",
                    pars = list(miss.thresh = 0.22, max.rf = 0.3))
```

A threshold of `miss.thresh = 0.22` ensures that all markers below that
proportion are restored. Recall that this mechanism assigns markers to the most
suitable linkage group but does not reconstruct the map.

It is worth re-examining the heat map at this point to establish whether the
restoration has been successful. The display is confined to linkage groups L.3,
L.5, L.8 and L.9, to determine whether the additional markers bear on the
possible merging of those groups.

```{r ex21, fig.width = 9.8, fig.height = 5.6, warning = FALSE, fig.cap = "Heat map of linkage groups L.3, L.5, L.8 and L.9 of mapBC6.", fig.alt = "Heat map of linkage groups L.3, L.5, L.8 and L.9 of mapBC6."}
heatMap(mapBC6, chr = c("L.3", "L.5", "L.8", "L.9"), lmax = 70)
```

Genuine linkage is evident between L.3 and L.5, and between L.8 and L.9. These
two pairs are merged with `mergeCross()` and the groups renamed, giving the
seven linkage groups expected for the barley genome.

```{r ex22}
mapBC6 <- mergeCross(mapBC6, merge = list("L.3" = c("L.3", "L.5"),
                                          "L.8" = c("L.8", "L.9")))
names(mapBC6$geno) <- paste("L.", 1:7, sep = "")
mapBC7 <- mstmap(mapBC6, bychr = TRUE, trace = FALSE, dist.fun = "kosambi",
                 p.value = 2)
chrlen(mapBC7)
```

Since the optimal number of linkage groups has been established, reconstruction
proceeds by linkage group alone, through `p.value = 2`. Anchoring is
unnecessary here, as the orientation of the groups has not yet been formally
determined.

The lengths of L.1, L.2 and L.4 are somewhat elevated, indicating excessive
recombination across those groups. The cause is apparent from the genotype
profiles.

```{r ex24, fig.width = 11.9, fig.height = 7, warning = FALSE, fig.cap = "For individual genotypes, the number of recombinations, double recombinations and missing values for mapBC7.", fig.alt = "For individual genotypes, the number of recombinations, double recombinations and missing values for mapBC7."}
pg1 <- profileGen(mapBC7, bychr = FALSE, stat.type = c("xo", "dxo", "miss"),
                  id = "Genotype", xo.lambda = 14, layout = c(1, 3), lty = 2,
                  cex = 0.7)
```

Introducing the markers with 10 to 20% missing values has exposed two further
problematic lines carrying a high proportion of missing values across the
genome. These are removed and the map reconstructed.

```{r ex25}
mapBC8 <- subsetCross(mapBC7, ind = !pg1$xo.lambda)
mapBC9 <- mstmap(mapBC8, bychr = TRUE, dist.fun = "kosambi", trace = FALSE,
                 p.value = 2)
chrlen(mapBC9)
```

Removing two lines does not affect the number of linkage groups, so
reconstruction is again by linkage group only. The lengths of most groups are
now appreciably reduced.

## Restoring the distorted markers

```{r ex27, fig.width = 11.9, fig.height = 7, warning = FALSE, fig.cap = "Marker profiles of the negative log10 p-value for the test of segregation distortion and the allele proportions for mapBC9.", fig.alt = "Marker profiles of segregation distortion and allele proportions for mapBC9."}
profileMark(mapBC9, stat.type = c("seg.dist", "prop"), layout = c(1, 3),
            type = "l")
```

A spike of segregation distortion appears on L.2 which does not appear to be
biological in origin and should be removed. Significant distortion is also
apparent on L.3, L.6 and L.7, indicating that some of the markers with 10 to
20% missing values restored earlier carry a degree of distortion themselves.
The marker positions on the horizontal axis suggest these regions are sparse.
The 295 markers held in the `"seg.distortion"` element may account for that
sparsity, and are restored to establish their effect.

```{r ex28}
dm <- markernames(mapBC9, "L.2")[statMark(mapBC9, chr = "L.2",
                                          stat.type = "marker")$marker$neglog10P > 6]
mapBC10 <- drop.markers(mapBC9, dm)
mapBC11 <- pushCross(mapBC10, type = "seg.distortion",
                     pars = list(seg.ratio = "70:30"))
mapBC12 <- mstmap(mapBC11, bychr = TRUE, trace = FALSE, dist.fun = "kosambi",
                  p.value = 2)
round(chrlen(mapBC12) - chrlen(mapBC9), 5)
nmar(mapBC12) - nmar(mapBC10)
```

A distortion ratio of 70:30 was chosen, after inspection of the table held in
the `"seg.distortion"` element, to ensure that all markers were restored. The
linkage group lengths change only negligibly from the previous map with the
distorted markers removed, which indicates that the additional markers have been
inserted successfully, nearly all of them into L.3, L.6 and L.7.

Finally the 39 co-locating markers held in the `"co.located"` element are
restored, each placed adjacent to the marker it co-locates with. Markers of this
type carry an immediate linkage group assignment.

```{r ex29}
mapBC <- pushCross(mapBC12, type = "co.located")
names(mapBC)
```

The additional elements have been removed from the object, leaving only
`"pheno"` and `"geno"`. The final map has the following structure.

| | L.1 | L.2 | L.3 | L.4 | L.5 | L.6 | L.7 | Total | Average |
|---|---|---|---|---|---|---|---|---|---|
| Number of markers | 681 | 593 | 335 | 679 | 233 | 279 | 219 | 3019 | 431 |
| Length (cM) | 225.9 | 224.4 | 157.6 | 205.5 | 195.1 | 145.5 | 142.9 | 1297.2 | 185.3 |
| Average interval (cM) | 0.33 | 0.39 | 0.48 | 0.31 | 0.84 | 0.53 | 0.66 | | 0.51 |

## Post-construction development

It is common to develop a linkage map further by inserting additional markers.
This may be a straightforward fine mapping exercise, in which the linkage group
of the additional markers is known in advance, or a more involved task such as
inserting markers from an older map into a newly constructed one.

Either can be problematic. External manipulation, such as the removal of
non-concurrent genotypes, may be required before insertion, and the map may need
partial or complete reconstruction afterwards. An unfortunate consequence of
reconstruction is the potential loss of known linkage group identities.

ASMap allows additional markers to be inserted while preserving those
identities. The methods below assume that the additional markers and the
constructed map are R/qtl cross objects of the same class. In each example the
additional markers are obtained by sampling from the final map, `mapBC`.

```{r add1}
set.seed(123)
add1 <- drop.markers(mapBC, markernames(mapBC)[sample(1:3019, 2700, replace = FALSE)])
mapBCs <- drop.markers(mapBC, markernames(add1))
add3 <- add2 <- add1
add2 <- subset(add2, chr = "L.1")
add3$geno[[1]]$data <- pull.geno(add1)
add3$geno[[1]]$map <- 1:ncol(add3$geno[[1]]$data)
names(add3$geno[[1]]$map) <- markernames(add1)
names(add3$geno)[1] <- "ALL"
add3 <- subset(add3, chr = "ALL")
```

### Combining two maps of the same population

Populations are frequently genotyped on more than one platform, with separate
maps constructed from each. A map built from an Illumina 90K SNP array, for
instance, may coexist with a legacy map built from SSR or DArT markers. The
`add1` object mimics such an older map of `mapBC`, with 319 markers spanning the
seven linkage groups. Since the linkage groups are known in both maps, it
remains only to combine them sensibly and reconstruct.

```{r add2}
add1 <- subset(add1, ind = 2:300)
full1 <- combineMap(mapBCs, add1, keep.all = TRUE)
full1 <- mstmap(full1, bychr = TRUE, trace = FALSE, anchor = TRUE, p.value = 2)
```

This is a characteristic application of
[`combineMap()`](../reference/combineMap.html), described in
[Manipulating linkage maps](manipulating-maps.html). The maps are first merged
on their matching genotypes; because the first genotype of `add1` has been
removed, missing cells are inserted in the first genotype of `mapBCs` for the
markers contributed by `add1`. The function recognises that both maps share
linkage group names and places the markers from shared groups together.
Reconstruction by linkage group with `p.value = 2` then retains those
identities, and `anchor = TRUE` preserves the orientation of the larger map.

### Fine mapping

In marker assisted selection it is common to increase marker density in a
specific genomic region, so that the position of quantitative trait loci may be
identified more accurately. The `add2` object contains markers from the L.1
group of `mapBCs` alone. Where the linkage group of the additional markers is
known in advance and matches a group in the constructed map, insertion proceeds
much as above.

```{r add3}
add2 <- subset(add2, ind = 2:300)
full2 <- combineMap(mapBCs, add2, keep.all = TRUE)
full2 <- mstmap(full2, chr = "L.1", bychr = TRUE, trace = FALSE, anchor = TRUE,
                p.value = 2)
```

The removal of the first genotype of `add2` again introduces missing values in
the first genotype of `full2`. The reconstruction differs only in that ordering
is required within L.1 alone.

### Unknown linkage groups

The linkage group of the additional markers is not always known in advance. An
incomplete marker set may have been used to construct the map, or a secondary
set may be available from an unconstructed map. Here the `add3` object consists
of a single group named `ALL` containing all the markers that spanned the seven
groups of `add1`.

```{r add4}
add3 <- subset(add3, ind = 2:300)
full3 <- combineMap(mapBCs, add3, keep.all = TRUE)
full3 <- pushCross(full3, type = "unlinked", unlinked.chr = "ALL")
full3 <- mstmap(full3, bychr = TRUE, trace = FALSE, anchor = TRUE, p.value = 2)
```

`combineMap()` again merges the maps, avoiding any need to match the dimensions
of the marker sets manually. `pushCross()` with `type = "unlinked"` and
`unlinked.chr = "ALL"` then distributes the additional markers among the
remaining linkage groups of `full3`, after which ordering within groups is all
that is required.

## Further reading

- [Worked example I](worked-example-construction.html) — the construction that
  precedes this article
- [Pulling and pushing markers](pulling-and-pushing.html) — the mechanism used
  throughout
- [Manipulating linkage maps](manipulating-maps.html) — `combineMap()` and
  `mergeCross()` in detail

## References
